---
title: "Data sources"

---
```{r}
#| include: false
options(
  CopernicusMarine_uid = Sys.getenv("COPERNICUS_USER"),
  CopernicusMarine_pwd = Sys.getenv("COPERNICUS_PWD")
)
```


**Needed libraries**:

```{r setup}

# Install pacman if needed
if (!require("pacman")) install.packages("pacman")

# A better way to load (and install, if needed) multiple pagages
pacman::p_load(tidyverse, sf, geodata,
               mregions2, leaflet, leaflet.extras2,
               marmap, terra, tidyterra, sdmpredictors,
               rnaturalearth, rnaturalearthdata,
               ggspatial, CFtime, ncmeta,
               stars,cubelyr)

```


# Where I can get data for my maps?

There is a lot of free spatial data available on the web. For example see:

<https://freegisdata.rtwilson.com/>.

Remember, Google is your friend. And (for now), the AI is too.


# Marine boundaries and administrative areas

mregions2 is a rOpenSci package that offers access to data from Marine Regions (https://marineregions.org/) in R, including:

1) The Marine Regions Gazetteer, a standardised list of marine place names with unique identifiers.   

2) Maritime Boundaries: including EEZs, maritime boundaries, and more.

```{r}
library (mregions2)

# An overview of data types available in the Gazetteer
gt <- gaz_types()

ns <- gaz_search("North Sea") # Search for entries matching "North Sea"

# Get the geometry for the North Sea (MRGID 2350)
north_sea <- gaz_search("North Sea") |>
  filter(MRGID == 2350) |>
  gaz_geometry()  # Add the geometry

ggplot() +
  geom_sf(data = north_sea, fill = "steelblue", 
          alpha = 0.4, colour = "navy") +
  theme_bw() +
  labs(title = "North Sea (IHO boundary)",
       caption = "Source: Marine Regions Gazetteer via mregions2")

```

Now let's access some of the Marine Regions data products.  Probably the more useful are the EEZs.

```{r}

# An overview of available Marine Regions data products
mrp_list

# Download EEZ boundaries for Sweden only
# Filter using a CQL expression on the sovereign1 attribute 
sweden_eez <- mrp_get("eez",
                       cql_filter = "sovereign1 = 'Sweden'")

ggplot() +
  geom_sf(data = sweden_eez, fill = "darkorange", alpha = 0.3) +
  theme_bw() +
  labs(title = "Sweden EEZ")
```

Here we will get the EEZ's that are located within the Barent's Sea:

```{r}
# Find the Barents Sea in the Gazetteer
bs <- gaz_search("Barents Sea") |>
  select(MRGID, placeType, preferredGazetteerName)

# MRGID 4247 is the IHO Sea Area entry for the Barents Sea
barents_poly <- gaz_search(4247) |> 
  gaz_geometry()

# Download all EEZs then clip to the Barents Sea polygon
all_eez <- mrp_get("eez",
                   cql_filter = "sovereign1 IN ('Norway', 'Russia')")

st_is_valid(all_eez) # Check for validity...

all_eez <- st_make_valid(all_eez) # ...and fix it

barents_eez <- st_intersection(all_eez, barents_poly)

ggplot() +
  geom_sf(data = barents_eez, 
          aes(fill = geoname), alpha = 0.3) +
  theme_bw() 
```

Some times we need to get the limits of EEZs as lines (rather than polygons):

```{r}
sweden_bounds <- mrp_get("eez_boundaries",
                          cql_filter = "sovereign1 = 'Sweden' OR sovereign2 = 'Sweden'")


ggplot() +
  geom_sf(data=sweden_bounds,
          aes(color=line_name))
```

You can use the `mrp_view` function to explore the data products interactively.

```{r}
mrp_view("eez")
```


# Administrative divisions
They can also be obtained from the GADM database using the `geodata` package.

We can use the name or the 3 letter ISO code of the country we want.

```{r}
library(geodata)

spain<- gadm(country = "Spain", level = 1, path = "./data") %>%
  st_as_sf()  # Because by default we get an object of SpatVector class

ggplot() +
  geom_sf(data=spain,aes(fill=NAME_1))

```

# Natural Earth

A very good source for mapping data is Natural Earth <http://www.naturalearthdata.com/>.  As we know already, it can be accessed directly from R using the **rnaturalearth** package.

The data available at three scales: large (lots of detail), intermediate, and small (few details).  The package has pre-downloaded maps, and a function (`ne_download`) to download any of the datasets in the site.

```{r}


# Map extent: Baltic Sea 
xlim <- c(9, 32)
ylim <- c(53, 66)

# Raster: cross-blended hypsometric tints 
# HYP_50M_SR_W = shaded relief + hypsometric tints + water colouring
# Returns a 3-band RGB SpatRaster
hypso <- ne_download(
  scale    = 50,
  type     = "HYP_50M_SR_W",
  category = "raster"
)

# Crop to the Baltic region to reduce memory and speed up plotting
hypso_crop <- crop(hypso, ext(xlim[1], xlim[2], ylim[1], ylim[2]))

# Vector layers
europe <- ne_countries(continent = "Europe", scale = "large")

rivers    <- ne_download(scale = "large",
                          type = "rivers_lake_centerlines",
                          category = "physical")

lakes     <- ne_download(scale = "large", type = "lakes",
                          category = "physical")



# Plot 
ggplot() +

  # RGB hypsometric tint raster — no colour scale needed, it is already coloured
  geom_spatraster_rgb(data = hypso_crop, maxcell = 5e6) +

  # Country borders
  geom_sf(data = europe, fill = NA, colour = "grey30", linewidth = 0.3) +

  # Rivers
  geom_sf(data = rivers, colour = "steelblue", linewidth = 0.4, alpha = 0.8) +
  # Lakes
  geom_sf(data = lakes, fill = "steelblue", colour = NA, alpha = 0.7) +
  coord_sf(xlim = xlim, ylim = ylim, expand = FALSE) +
  theme_bw() +
  theme(legend.position = "none") +
  labs(
    title   = "The Baltic Sea region",
    x = NULL, y = NULL
  )


```


# Bathymetry

We also have shown how to use the `marmap` package to access bathymetry data from the ETOPO1 Global Relief Model (<https://www.ngdc.noaa.gov/mgg/global/>).


```{r}

xlim <- c(-5, 10)
ylim <- c(51, 62)

europe <- ne_countries(continent = "Europe", scale = "large")

bathy <- getNOAA.bathy(
  lon1 = xlim[1], lon2 = xlim[2],
  lat1 = ylim[1], lat2 = ylim[2],
  resolution = 2, keep = TRUE, path = "./data") |>
  marmap::as.raster() |>   # returns a raster object
  rast()   

bathy[bathy>= 0] <- NA   # keep only depths

# Clamp very deep locations 
bathy <- clamp(bathy, lower = -1500, upper = 0, values = FALSE) 

ggplot() +

  geom_spatraster(data = bathy) +

  scale_fill_distiller(
  palette  = "Blues",
  direction = -1,          # dark = deep, light = shallow
  na.value = NA,
  name     = "Depth (m)",
  limits   = c(-1500, 0))+
  geom_spatraster_contour(data = bathy,
               breaks    = c(-50,-100,-200, -500, -1000, -2000),
               colour    = "gray50", linewidth = 0.2, alpha = 0.5) +
  geom_sf(data=europe, fill="gray90", color="gray90") +
  coord_sf(xlim = xlim, ylim = ylim, expand = FALSE) +
  theme_bw() +
  theme(legend.position = "right") +
  labs(title   = "North Sea",
       x = NULL, y = NULL)

```
`marmap `also can read files from the GEBCO bathymetry database (but you have to download
  them manually) from <https://www.gebco.net/>.  If you do this, select the GeoTiff option so you
  can read them directly into a raster object with `terra`.

`ggoceanmaps `is another package that can be used to download and plot bathymetry data.



# sdmpredictors
The package `sdmpredictors` provides access to two environmental datasets:

1. Bio-Oracle (<https://www.bio-oracle.org/>), a set of rasters with geophysical, biotic and environmental data on the surface and the seabed. Includes many parameters: temperature, chlorophyll concentration, dissolved oxygen, etc.

2. MARSPEC (<https://marspec.org/>) is a set of climatic and geophyisical layers focusing on pelagic and shallow waters.


```{r}
layers <- list_layers(datasets = "Bio-ORACLE", marine = TRUE)

# Near-bottom layers have "_bd" in the layer code
bottom_layers <- layers |>
  filter(str_detect(layer_code, "_bd")) |>
  select(layer_code, name, units, description)

# Set a persistent directory for dowloading
options(sdmpredictors_datadir = "./data/sdmpredictors")

env_bottom <- load_layers(
  layercodes = c(
    "BO2_tempmean_bdmean",      # mean near-bottom temperature (°C)
    "BO2_salinitymean_bdmean",  # mean near-bottom salinity (PSS)
    "BO2_dissoxmean_bdmean"     # mean near-bottom dissolved oxygen (ml/l)
  ),
  equalarea = FALSE) |>    # keep WGS84 geographic CRS
rast()

# Rename layers for clarity
names(env_bottom) <- c("Temperature",
                        "Salinity",
                        "Dissolved oxygen")
                       
xlim <- c(-5, 30)
ylim <- c(48, 65)

env_bottom_crop <- crop(env_bottom, ext(xlim[1], xlim[2], ylim[1], ylim[2]))
                       

p1 <- ggplot() +
  geom_spatraster(data = env_bottom_crop[[1]]) +
  scale_fill_viridis_c(option = "viridis", na.value = NA, name = NULL) +
  coord_sf(xlim = xlim, ylim = ylim, expand = FALSE) +
  theme_void() +
  theme(legend.position = "bottom")+
  labs(title="Temperature (°C)")
        
p2 <- ggplot() +
  geom_spatraster(data = env_bottom_crop[[2]]) +
  scale_fill_viridis_c(option = "viridis", na.value = NA, name = NULL) +
  coord_sf(xlim = xlim, ylim = ylim, expand = FALSE) +
  theme_void() +
  theme(legend.position = "bottom") +
  labs(title="Salinity (PPS)")

p3 <- ggplot() +
  geom_spatraster(data = env_bottom_crop[[3]]) +
  scale_fill_viridis_c(option = "viridis", na.value = NA, name = NULL) +
  coord_sf(xlim = xlim, ylim = ylim, expand = FALSE) +
  theme_void() +
  theme(legend.position = "bottom") +
  labs(title="Dissolved oxygen (ml/l)")

library(patchwork)
p1+p2+p3


```
# Copernicus Marine ... and showing off the star package

To download data from Copernicus Marine, you can use the `CopernicusMarine` package.  You can install it like this:

```{r}
#| eval: false
install.packages("CopernicusMarine",
                 repos = c("https://pepijn-devries.r-universe.dev",
                           "https://cloud.r-project.org"))
```

You will need to create an account in Copernicus Marine before you can get data.  You can put them in the .Renviron file, like this:

COPERNICUS_USER=your_username
COPERNICUS_PWD=your_password

And put this in your .Rprofile file:

options(
  CopernicusMarine_uid = Sys.getenv("COPERNICUS_USER"),
  CopernicusMarine_pwd = Sys.getenv("COPERNICUS_PWD")
)

So the username and passwords get loaded every time you start R.

You can also put them in your code directly using `options`, but you will have to do this in every R session:

```{r}
#| eval: false
options(
  CopernicusMarine_uid = "your_username",
  CopernicusMarine_pwd = "your_password"
)
```

With this, we can download some data.  Let's get sea surface temperature from the Global Phisics Analysis and Forecast.  First some exploration to select which data to dowload.

```{r}
library(CopernicusMarine)

# List all products matching a search term
products <- cms_products_list(contain = "GLOBAL_ANALYSISFORECAST_PHY")

layers <- cms_product_services("GLOBAL_ANALYSISFORECAST_PHY_001_024")

# Get names of variables in the product
details <- cms_product_details("GLOBAL_ANALYSISFORECAST_PHY_001_024")

variables <- details$properties$allVariables |>
  unlist() 

cms_cite_product("GLOBAL_ANALYSISFORECAST_PHY_001_024")$doi

dt <- cms_product_details("GLOBAL_ANALYSISFORECAST_PHY_001_024") 
dt <- dt$properties

region <- c(-5, 51, 10, 62)  # xmin, ymin, xmax, ymax

```



```{r}
#| eval: false
# Download surface temperature between 0 and 20m

sst <- cms_download_subset(
  product       = "GLOBAL_ANALYSISFORECAST_PHY_001_024",
  layer         = "cmems_mod_glo_phy-thetao_anfc_0.083deg_P1D-m",
  variable      = "thetao",
  region        = c(-5, 51, 10, 62),        # xmin, ymin, xmax, ymax
  timerange     = c("2024-08-01", "2024-08-31"),
  verticalrange = c(0, -20)              # surface layer

)

write_rds(sst, "./data/north_sea_sst_aug2024.rds")

```

Now we can download the data with the `cms_download_subset` function.  Note that the data is dowloaded as a `stars` object.

```{r}

sst <- read_rds("./data/north_sea_sst_aug2024.rds")

# Check how many depth levels were returned
st_get_dimension_values(sst, "elevation")

# --- Inspect the time dimension 
st_get_dimension_values(sst, "time")

# Mean over time only — keeps spatial + elevation dimensions
sst_mean <- st_apply(sst, c("longitude", "latitude", "elevation"),
                     mean, na.rm = TRUE)

# Extract surface and deepest level
n_elev <- dim(sst_mean)["elevation"]

sst_mean_surface <- dplyr::slice(sst_mean, along = "elevation", index = 1)
sst_mean_deep    <- dplyr::slice(sst_mean, along = "elevation", index = n_elev)


p1 <- ggplot() +
  geom_stars(data = sst_mean_surface) +
  scale_fill_viridis_c(option = "inferno", name = "SST (°C)",
                       na.value = NA, limits = c(10, 23)) +
  coord_sf() +
  theme_void() +
  labs(title = "Surface (~0.5 m)")

p2 <- ggplot() +
  geom_stars(data = sst_mean_deep) +
  scale_fill_viridis_c(option = "inferno", name = "SST (°C)",
                       na.value = NA, limits = c(10, 23)) +
  coord_sf() +
  theme_void() +
  labs(title = "Surface (~50 m)")

p1 + p2

# --- Extract a single day 
sst_day1 <- filter(sst, time == as.Date("2024-08-01"))

# stars objects can be subsetted as [attribute, x, y, elevation, time]
sst_day1 <- sst[,,,, 1]   # all x, y, elevation, first time step

# Plot a single day
ggplot() +
  geom_stars(data = sst_day1) +
  scale_fill_viridis_c(option = "inferno", name = "SST (°C)", na.value = NA) +
  coord_sf() +
  theme_void() +
  labs(title = "North Sea SST — 1 August 2024",
       caption = "Source: E.U. Copernicus Marine Service")

# Convert the monthly mean to terra SpatRaster if preferred 
#sst_rast <- rast(sst_mean)

```



# On a tangent... curvilinear grids with stars!
```{r}
prec_file <- system.file("nc/test_stageiv_xyt.nc", package = "stars")

# curvilinear = c("lon", "lat") tells stars which variables in the file
# hold the 2D coordinate arrays — each cell has its own lon/lat pair
prec <- read_ncdf(prec_file,
                  curvilinear  = c("lon", "lat"),
                  ignore_bounds = TRUE)

# Take the first time step
prec_t1 <- dplyr::slice(prec, along = "time", index = 1)

# Plot
ggplot() +
  geom_stars(data = prec_t1) +
  scale_fill_viridis_c(option = "mako", direction = -1,
                       na.value = NA, name = "Precip\n(kg/m²)",
                       trans = "sqrt") +
  coord_sf() +
  theme_void() +
  labs(title   = "Hourly precipitation — curvilinear grid",
       caption = "Source: stars built-in example data")

```



# EMODNET
https://emodnet.github.io/emodnet.wfs/index.html


```{r}

# Install if needed
pak::pak("EMODnet/emodnet.wfs") # Modern way to install packages from GitHub

library(emodnet.wfs)

# Suppress the "WFS client created" messages
options("emodnet.wfs.quiet" = TRUE)

# Step 1: see what services are available 
services <- emodnet_wfs() |>
  select(emodnet_thematic_lot, service_name)

# Step 2: initialise a client for human activities 
wfs <- emodnet_init_wfs_client(service = "human_activities")

# Step 3: list available layers... 
layers_info <- emodnet_get_wfs_info(wfs)

# ...and search for wind-related layers in the list
layers_info |>
  filter(str_detect(layer_name, "wind")) |>
  select(layer_name, title)

# Step 4: download a layer — offshore wind farms
wind_farms <- emodnet_get_layers(
  wfs    = wfs,
  layers = "windfarms"   # or whatever the correct name is
)

wind_sf <- wind_farms$windfarms

# Step 5: crop to the North Sea and plot --------------------------------
xlim <- c(-5, 10)
ylim <- c(51, 58)

wind_ns <- wind_sf |>
  st_crop(xmin = xlim[1], xmax = xlim[2],
          ymin = ylim[1], ymax = ylim[2])

ggplot() +
  geom_sf(data=wind_ns, aes(color = status), alpha = 0.7) +
  geom_sf(data=europe, fill="gray70")+
  scale_fill_brewer(palette = "Set2", name = "Status") +
  coord_sf(xlim = xlim, ylim = ylim, expand = FALSE) +
  theme_void() +
  theme(legend.position = "bottom") +
  labs(
    title   = "Offshore wind farms — North Sea",
    caption = "Source: EMODnet Human Activities via emodnet.wfs"
  )
```


