---
title: "Mapping with ggplot2"
execute:
  enabled: false
---

# Getting started

The [ggplot2 book](https://ggplot2-book.org) is the definitive reference for everything covered here and much more.

## Libraries

```{r setup}
library(tidyverse)          # ggplot2, dplyr, readr, etc.
library(sf)                 # spatial objects and geom_sf
library(rnaturalearth)      # modern basemap data as sf objects
library(rnaturalearthdata)  # data dependency for rnaturalearth
library(ggspatial)          # scale bars, north arrows
library(patchwork)          # combining multiple plots
library(tidyterra)          # ggplot2 integration for terra rasters
library(terra)              # raster data
library(marmap)             # bathymetry data
```

## Key components of ggplot2

**Data used**:
```{r}
minke <- read_csv("https://media.githubusercontent.com/media/jmburgos/TCMSAR26/refs/heads/main/data/minke.csv")
glimpse(minke)
```

ggplot2 has **three** key components:

1. A **data** argument — a `data.frame` or derivative (`tibble`, `sf`, ...)
2. A set of **aesthetic mappings** via `aes()` — connecting variables to visual properties
3. At least one **layer** via a `geom_*()` function — describing how to render observations

```{r}
ggplot(data = minke) +
  aes(x = lon, y = lat, colour = sex) +
  geom_point()
```

Different syntax, equivalent outcome — all four lines below produce the same plot:

```{r eval=FALSE}
ggplot()                                   + geom_point(data = minke, aes(lon, lat, colour = sex))
ggplot(data = minke)                       + geom_point(aes(lon, lat, colour = sex))
ggplot(minke)                              + geom_point(aes(lon, lat, colour = sex))
ggplot(minke, aes(lon, lat, colour = sex)) + geom_point()
```

A ggplot can be stored as an object for later use and built upon:

```{r}
p <- ggplot(minke) + geom_point(aes(lon, lat, colour = sex))
p
```

This plot maps longitude to x and latitude to y, but two things are missing: a background reference map and a proper map projection.

---

# The modern approach: `geom_sf()`

The best way to make maps in ggplot2 is with `geom_sf()`. It works directly with `sf` objects, automatically handles coordinate reference systems via `coord_sf()`, and eliminates the need to manage projections manually.

## Basemaps with `rnaturalearth`

The **rnaturalearth** package provides country polygons and coastlines as `sf` objects at multiple resolutions — a modern, well-maintained replacement for the `maps`/`mapdata` packages.

```{r}
# Country polygon at 1:50m resolution
iceland_sf <- ne_countries(country = "Iceland", scale = "medium",
                           returnclass = "sf")

ggplot() +
  geom_sf(data = iceland_sf, fill = "grey80", colour = "white") +
  theme_bw()
```

Now add the minke whale observations on top. First convert the point data to an `sf` object:

```{r}
minke_sf <- minke %>%
  st_as_sf(coords = c("lon", "lat"), crs = 4326)

ggplot() +
  geom_sf(data = iceland_sf, fill = "grey80") +
  geom_sf(data = minke_sf, aes(colour = sex), size = 2, alpha = 0.7) +
  scale_colour_brewer(name = "Sex", palette = "Set1") +
  labs(x = NULL, y = NULL) +
  theme_bw()
```

## Projections with `coord_sf()`

`coord_sf()` controls the CRS used when rendering all map layers. Every layer is automatically reprojected to match — you never need to reproject your data manually just for plotting purposes.

```{r}
p_base <- ggplot() +
  geom_sf(data = iceland_sf, fill = "grey80") +
  geom_sf(data = minke_sf, aes(colour = sex), alpha = 0.7) +
  scale_colour_brewer(name = "Sex", palette = "Set1") +
  scale_x_continuous(NULL, NULL) +
  scale_y_continuous(NULL, NULL)

# WGS84 geographic (default)
p1 <- p_base + coord_sf(crs = 4326) + labs(title = "WGS84")

# Lambert Azimuthal Equal Area centred on Iceland
p2 <- p_base +
  coord_sf(crs = "+proj=laea +lat_0=65 +lon_0=-19") +
  labs(title = "LAEA")

# Stereographic
p3 <- p_base +
  coord_sf(crs = "+proj=stere +lat_0=90 +lat_ts=65 +lon_0=-19") +
  labs(title = "Stereographic")

p1 + p2 + p3
```

::: {.callout-important}
`coord_map()` and `coord_quickmap()` are **deprecated** as of ggplot2 3.5.0. Use `coord_sf()` for all map projections. It handles true cartographic projections correctly and integrates seamlessly with `geom_sf()`.
:::

## Zooming in

Pass `xlim` and `ylim` to `coord_sf()` to restrict the visible area. The values are in the units of the active CRS (degrees for geographic coordinates):

```{r}
world_sf <- ne_countries(scale = "medium", returnclass = "sf")

ggplot() +
  geom_sf(data = world_sf, fill = "grey80") +
  geom_sf(data = minke_sf, aes(colour = sex), alpha = 0.8) +
  coord_sf(xlim = c(-26, -12), ylim = c(62.5, 67.5)) +
  scale_colour_brewer(name = "Sex", palette = "Set1") +
  labs(x = NULL, y = NULL) +
  theme_bw()
```

## Labels with `geom_sf_text()` and `geom_sf_label()`

```{r}
nfu <- read_sf("data/nephrops_fu.gpkg")
nfu_centroids <- st_centroid(nfu)

ggplot() +
  geom_sf(data = nfu, fill = NA, colour = "steelblue") +
  geom_sf_label(data = nfu_centroids, aes(label = name),
                size = 2.5, label.padding = unit(0.15, "lines")) +
  theme_bw()
```

---

# The legacy approach: `geom_polygon()` with `map_data()`

Before `geom_sf()` existed, background maps were built from data frames of coordinates drawn with `geom_polygon()`. You will still encounter this in older code, so it is worth understanding.

```{r}
library(maps)
library(mapdata)

iceland_df <- map_data("worldHires", region = "Iceland")
glimpse(iceland_df)
```

This is a data frame of longitude/latitude coordinates with a `group` column identifying which polygon each point belongs to. Plotting with different geoms illustrates why `group` and point order matter:

```{r fig.height=6}
m <- ggplot(iceland_df, aes(long, lat, group = group))

(m + geom_point()   + labs(title = "geom_point")) +
(m + geom_line()    + labs(title = "geom_line")) +
(m + geom_path()    + labs(title = "geom_path")) +
(m + geom_polygon() + labs(title = "geom_polygon"))
```

Only `geom_path()` and `geom_polygon()` produce sensible maps because they respect coordinate **order** within each group.

::: {.callout-tip}
Prefer `geom_sf()` for new code. The `geom_polygon()` / `map_data()` approach requires manual management of the `group` variable, does not support reprojection, and can produce distorted aspect ratios.
:::

---

# More layers with survey data

```{r}
smb <- read_csv("data/smb_summary.csv")
smb2019 <- filter(smb, year == 2019)

# A reusable basemap — build it once, add layers as needed
basemap <- ggplot() +
  geom_sf(data = iceland_sf, fill = "grey80", colour = "white") +
  coord_sf(xlim = c(-27, -11), ylim = c(62.5, 67.5)) +
  scale_x_continuous(NULL, NULL) +
  scale_y_continuous(NULL, NULL) +
  theme_bw()
```

## `geom_segment()` — tow tracks

`geom_segment()` draws straight lines between start and end coordinates, ideal for tow tracks:

```{r}
basemap +
  geom_segment(data = smb2019,
               aes(x = lon1, y = lat1, xend = lon2, yend = lat2),
               linewidth = 0.3)
```

## `geom_point()` — bubble plots

Mapping a continuous variable to `size` creates a bubble chart. Use `scale_size_area()` so that **area** (not radius) is proportional to the value, and zero maps to zero area:

```{r}
basemap +
  geom_point(data = smb2019,
             aes(lon1, lat1, size = cod_n),
             alpha = 0.3, colour = "red") +
  scale_size_area(max_size = 20, name = "Cod (n)")
```

::: {.panel .panel-warning}
::: panel-heading
Exercise
:::
::: panel-body
Recreate the bubble plot above for a different species. Add `geom_segment()` for tow tracks underneath the bubbles, so the tracks appear behind the points.
:::
:::

## `geom_path()` — vessel tracks

`geom_path()` connects points **in the order they appear in the data**, making it ideal for time-ordered GPS tracks:

```{r}
#| eval: false
#track <- read_csv("data/is_smb_vms2019.csv")
#track <- read_csv("data/small_vms.csv")

basemap +
  geom_path(data = track,
            aes(x = lon, y = lat, colour = factor(id)),
            linewidth = 0.4, alpha = 0.8) +
  scale_colour_brewer(name = "Vessel", palette = "Set1")
```

## `geom_tile()` — gridded summaries

`geom_tile()` draws rectangles centred on each (x, y) coordinate, useful for pre-gridded data. Note the layer order: draw tiles first, then the coastline on top:

```{r}
#| eval: false
s <- smb2019 %>%
  group_by(ir_lon, ir_lat) %>%
  summarise(Cod = mean(cod_n), .groups = "drop")

ggplot() +
  geom_tile(data = s, aes(x = ir_lon, y = ir_lat, fill = Cod)) +
  geom_sf(data = iceland_sf, fill = "grey60") +
  coord_sf(xlim = c(-27, -11), ylim = c(62.5, 67.5)) +
  scale_fill_viridis_c(option = "B", direction = -1, name = "Mean cod (n)") +
  scale_x_continuous(NULL, NULL) +
  scale_y_continuous(NULL, NULL) +
  theme_bw()
```

::: {.panel .panel-warning}
::: panel-heading
Exercise
:::
::: panel-body
Repeat the tile plot above but use the `grade()` function to bin at double the resolution of ICES rectangles (0.5° longitude × 0.25° latitude):

```{r}
#| eval: false
grade <- function(x, dx) {
  brks <- seq(floor(min(x)), ceiling(max(x)), dx)
  ints <- findInterval(x, brks, all.inside = TRUE)
  (brks[ints] + brks[ints + 1]) / 2
}
```
:::
:::

## `geom_raster()` and `geom_contour()` — bathymetry

For regular raster grids (equal-resolution cells), `geom_raster()` is faster than `geom_tile()`. Here we use `marmap` to download ETOPO1 bathymetry data:

```{r}
xlim <- c(-28, -10)
ylim <- c(62.5, 67.5)
depth <- 
  getNOAA.bathy(lon1 = xlim[1], lon2 = xlim[2],
                lat1 = ylim[1], lat2 = ylim[2],
                resolution = 1) |> 
  fortify()  # turn the object into a data.frame
glimpse(depth)

#depth_df <- as.data.frame(depth_bathy, xy = TRUE) %>%
#  rename(x = lon, y = lat, z = depth) %>%
#  filter(z <= 0)

ggplot() +
  geom_raster(data = depth, aes(x = x, y = y, fill = z)) +
  geom_sf(data = iceland_sf, fill = "grey70") +
  scale_fill_viridis_c(option = "D", name = "Depth (m)") +
  coord_sf(xlim = xlim, ylim = ylim) +
  scale_x_continuous(NULL, NULL) +
  scale_y_continuous(NULL, NULL) +
  theme_bw()
```

Add depth contour lines with `geom_contour()`:

```{r}
ggplot() +
  geom_raster(data = depth, aes(x = x, y = y, fill = z)) +
  geom_contour(data = depth,
               aes(x = x, y = y, z = z),
               breaks = c(-50, -100, -200, -400, -1000),
               linewidth = 0.3, colour = "white") +
  geom_sf(data = iceland_sf, fill = "grey70") +
  scale_fill_viridis_c(option = "D", name = "Depth (m)") +
  coord_sf(xlim = xlim, ylim = ylim) +
  scale_x_continuous(NULL, NULL) +
  scale_y_continuous(NULL, NULL) +
  theme_bw()
```

### Plotting `terra` rasters directly with `tidyterra`

The **tidyterra** package provides `geom_spatraster()`, which accepts `SpatRaster` objects directly — no manual conversion to a data frame needed:

```{r}
mintemp <- rast("data/Iceland_minbtemp.tif")

ggplot() +
  geom_spatraster(data = mintemp) +
  geom_sf(data = iceland_sf, fill = "grey60") +
  scale_fill_viridis_c(option = "C", name = "Min bottom\ntemp (°C)",
                       na.value = "transparent") +
  coord_sf() +
  theme_bw()
```

::: {.panel .panel-warning}
::: panel-heading
Exercise
:::
::: panel-body
1. Create a depth raster map of a region of your choice using `getNOAA.bathy()`.
2. Add contour lines at depths that are ecologically meaningful for your region.
3. Overlay survey station points on top, sized by a species of your choice.
:::
:::

---

# Facetting

Facetting splits data by one or two variables and draws the same plot for each subset — very effective for showing temporal or categorical patterns.

## `facet_wrap()`

```{r}
smb_sel <- filter(smb, year %in% c(1995, 2000, 2005, 2010, 2015, 2019))

ggplot() +
  geom_sf(data = iceland_sf, fill = "grey80") +
  geom_point(data = smb_sel,
             aes(lon1, lat1, size = haddock_kg),
             alpha = 0.15, colour = "red") +
  scale_size_area(max_size = 12, name = "Haddock (kg)") +
  coord_sf(xlim = c(-27, -11), ylim = c(62.5, 67.5)) +
  scale_x_continuous(NULL, NULL) +
  scale_y_continuous(NULL, NULL) +
  facet_wrap(~ year) +
  theme_bw()
```

## `facet_grid()`

`facet_grid()` creates a matrix of panels defined by two variables. Here, cod abundance is shown by year class (columns) and age (rows):

```{r fig.height=7}
rbya <- read_csv("data/is_smb_cod_rbya.csv") %>%
  filter(yc %in% c(1984, 1985, 1996, 1997))

ggplot() +
  geom_sf(data = iceland_sf, fill = "grey80") +
  geom_point(data = rbya,
             aes(lon1, lat1, size = n),
             alpha = 0.25, colour = "red") +
  scale_size_area(max_size = 8, name = "per 4 miles") +
  coord_sf(xlim = c(-27, -11), ylim = c(62.5, 67.5)) +
  scale_x_continuous(NULL, NULL) +
  scale_y_continuous(NULL, NULL) +
  facet_grid(age ~ yc) +
  labs(title = "Abundance of cod",
       subtitle = "Year class (columns) × age (rows)") +
  theme_bw()
```

In `facet_grid(rows ~ cols)`, the first variable defines rows and the second defines columns.

---

# Colour scales

## Continuous data: viridis palettes

The **viridis** palettes (`"viridis"`, `"magma"`, `"plasma"`, `"inferno"`, `"cividis"`, `"mako"`, `"rocket"`, `"turbo"`) are built into ggplot2. They are perceptually uniform, readable in greyscale, and accessible to colour-blind readers:

```{r}
p_tile <- ggplot() +
  geom_tile(data = s, aes(x = ir_lon, y = ir_lat, fill = Cod)) +
  geom_sf(data = iceland_sf, fill = "grey60") +
  coord_sf(xlim = c(-27, -11), ylim = c(62.5, 67.5)) +
  scale_x_continuous(NULL, NULL) +
  scale_y_continuous(NULL, NULL) +
  theme_bw()

p1 <- p_tile + scale_fill_viridis_c(option = "D", name = "viridis")
p2 <- p_tile + scale_fill_viridis_c(option = "B", direction = -1, name = "magma (rev)")
p3 <- p_tile + scale_fill_viridis_c(option = "C", name = "plasma")

p1 + p2 + p3
```

## Colour scale transformations

When data are highly skewed, transforming the colour scale can reveal spatial patterns more clearly. The `transform` argument (renamed from `trans` in ggplot2 3.5.0) accepts standard transformation names:

```{r}
p_log  <- p_tile + scale_fill_viridis_c(option = "B", direction = -1,
                                         transform = "log1p",
                                         name = "log1p scale")
p_sqrt <- p_tile + scale_fill_viridis_c(option = "B", direction = -1,
                                         transform = "sqrt",
                                         name = "sqrt scale")
p_log + p_sqrt
```

## Discrete intervals with `cut()`

Binning continuous data into labelled classes can be clearer than a smooth gradient, especially for reporting:

```{r}
s_cut <- s %>%
  mutate(Cod_class = cut(Cod,
                         breaks = c(0, 23, 62, 145, Inf),
                         labels = c("Low", "Medium", "High", "Very high"),
                         include.lowest = TRUE))

ggplot() +
  geom_tile(data = s_cut, aes(x = ir_lon, y = ir_lat, fill = Cod_class)) +
  geom_sf(data = iceland_sf, fill = "grey60") +
  coord_sf(xlim = c(-27, -11), ylim = c(62.5, 67.5)) +
  scale_fill_viridis_d(option = "B", direction = -1,
                       name = "Cod abundance",
                       na.value = "white") +
  scale_x_continuous(NULL, NULL) +
  scale_y_continuous(NULL, NULL) +
  theme_bw()
```

---

# Axis formatting

## Geographic degree labels

The helper functions below produce properly formatted degree–direction axis labels:

```{r}
scale_longitude <- function(min = -180, max = 180, step = 1, ...) {
  breaks <- seq(min, max, step)
  labels <- ifelse(breaks < 0,
                   paste0(abs(breaks), "\u00B0W"),
                   paste0(breaks, "\u00B0E"))
  scale_x_continuous(name = NULL, breaks = breaks, labels = labels, ...)
}

scale_latitude <- function(min = -90, max = 90, step = 0.5, ...) {
  breaks <- seq(min, max, step)
  labels <- ifelse(breaks < 0,
                   paste0(abs(breaks), "\u00B0S"),
                   paste0(breaks, "\u00B0N"))
  scale_y_continuous(name = NULL, breaks = breaks, labels = labels, ...)
}

ggplot() +
  geom_sf(data = iceland_sf, fill = "grey80") +
  coord_sf(xlim = c(-25, -12), ylim = c(63, 67)) +
  scale_longitude(min = -25, max = -12, step = 2) +
  scale_latitude(min = 63, max = 67, step = 1) +
  theme_bw()
```

## ICES statistical rectangle labels

ICES rectangles use a notation like "40G1" rather than decimal degrees. The functions below relabel axes accordingly. They require the **geo** package (`remotes::install_github("hafro/geo")`):

```{r eval=FALSE}
library(geo)

scale_longitude_ices <- function(min = -44, max = 68.5, step = 1, ...) {
  breaks <- seq(min + 0.5, max - 0.5, step)
  labels <- geo::d2ir(60, breaks) %>% stringr::str_sub(3)
  scale_x_continuous(name = NULL, breaks = breaks, labels = labels, ...)
}

scale_latitude_ices <- function(min = 36, max = 84.5, step = 0.5, ...) {
  breaks <- seq(min + 0.25, max - 0.25, step)
  labels <- geo::d2ir(breaks, 0) %>% stringr::str_sub(1, 2)
  scale_y_continuous(name = NULL, breaks = breaks, labels = labels, ...)
}

ggplot() +
  geom_sf(data = iceland_sf, fill = "grey80") +
  coord_sf(xlim = c(-25, -12), ylim = c(63, 67)) +
  scale_longitude_ices() +
  scale_latitude_ices() +
  theme(panel.grid.major = element_blank(),
        panel.grid.minor = element_line(linewidth = 0.5),
        axis.ticks = element_blank()) +
  theme_bw()
```

---

# Map embellishments

## Scale bars and north arrows with `ggspatial`

```{r}
ggplot() +
  geom_sf(data = iceland_sf, fill = "grey80") +
  annotation_scale(location = "bl", width_hint = 0.4) +
  annotation_north_arrow(
    location = "bl", which_north = "true",
    pad_x = unit(0.6, "in"), pad_y = unit(0.4, "in"),
    style = north_arrow_fancy_orienteering
  ) +
  coord_sf() +
  theme_bw()
```

## Themes for maps

`theme_void()` removes axes, grid lines, and background — often the cleanest choice for publication maps that carry their own spatial reference:

```{r}
ggplot() +
  geom_sf(data = iceland_sf, fill = "grey85", colour = "white") +
  geom_sf(data = minke_sf, aes(colour = sex), size = 1.8, alpha = 0.8) +
  scale_colour_brewer(name = "Sex", palette = "Set1") +
  annotation_scale(location = "br") +
  annotation_north_arrow(
    location = "tl",
    style = north_arrow_minimal,
    height = unit(0.8, "cm"), width = unit(0.8, "cm")
  ) +
  coord_sf() +
  theme_void() +
  theme(legend.position = "bottom",
        plot.title = element_text(face = "bold")) +
  labs(title = "Minke whale sightings", caption = "Source: MFRI")
```

---

# Combining plots with `patchwork`

**patchwork** provides an intuitive syntax for assembling ggplot objects. The `+` operator places plots side by side; `/` stacks them vertically:

```{r}
p_map <- ggplot() +
  geom_sf(data = iceland_sf, fill = "grey80") +
  geom_sf(data = minke_sf, aes(colour = sex), alpha = 0.7) +
  scale_colour_brewer(name = "Sex", palette = "Set1") +
  coord_sf() + theme_bw() + labs(title = "Location")

p_hist <- ggplot(minke, aes(x = length, fill = sex)) +
  geom_histogram(bins = 25, position = "identity", alpha = 0.6) +
  scale_fill_brewer(name = "Sex", palette = "Set1") +
  labs(x = "Length (cm)", y = "Count", title = "Length distribution") +
  theme_bw()

# Side by side, shared legend
(p_map + p_hist) + plot_layout(guides = "collect")
```

## Inset maps with `inset_element()`

`inset_element()` places one plot inside another — useful for a small context map within a detailed regional map:

```{r}
context_map <- ggplot() +
  geom_sf(data = world_sf, fill = "grey80", colour = "white",
          linewidth = 0.2) +
  geom_rect(aes(xmin = -27, xmax = -11, ymin = 62.5, ymax = 67.5),
            fill = NA, colour = "red", linewidth = 1) +
  coord_sf(xlim = c(-40, 30), ylim = c(45, 80)) +
  theme_void()

detail_map <- ggplot() +
  geom_sf(data = iceland_sf, fill = "grey80") +
  geom_sf(data = minke_sf, aes(colour = sex), alpha = 0.8) +
  scale_colour_brewer(name = "Sex", palette = "Set1") +
  coord_sf() + theme_bw() +
  labs(title = "Minke whale sightings")

detail_map +
  inset_element(context_map,
                left = 0.65, right = 1.0,
                bottom = 0.0, top = 0.35)
```

---

# Gridding point data

## Using `grade()` for custom grids

The `grade()` function bins coordinates to a regular grid and returns the midpoint of each cell — useful for summarising survey data at a chosen resolution:

```{r}
grade <- function(x, dx) {
  if (dx > 1) warning("Not tested for grids larger than one degree")
  brks <- seq(floor(min(x)), ceiling(max(x)), dx)
  ints <- findInterval(x, brks, all.inside = TRUE)
  (brks[ints] + brks[ints + 1]) / 2
}

smb2019_gridded <- smb2019 %>%
  mutate(glon = grade(lon1, 0.5),
         glat = grade(lat1, 0.25)) %>%
  group_by(glon, glat) %>%
  summarise(Cod = mean(cod_n), .groups = "drop")

ggplot() +
  geom_tile(data = smb2019_gridded,
            aes(x = glon, y = glat, fill = Cod)) +
  geom_sf(data = iceland_sf, fill = "grey60") +
  coord_sf(xlim = c(-27, -11), ylim = c(62.5, 67.5)) +
  scale_fill_viridis_c(option = "B", direction = -1,
                       transform = "log1p",
                       name = "Mean cod (n)") +
  scale_x_continuous(NULL, NULL) +
  scale_y_continuous(NULL, NULL) +
  theme_bw()
```

## Using `st_make_grid()` for spatial polygon grids

For a more rigorous approach that produces proper `sf` polygon grids respecting the CRS, use `st_make_grid()` then join with a spatial join:

```{r}
bbox_sfc <- st_bbox(c(xmin = -27, xmax = -11,
                      ymin = 62.5, ymax = 67.5),
                    crs = 4326) %>%
  st_as_sfc()

grid_sf <- st_make_grid(bbox_sfc,
                        cellsize = c(0.5, 0.25),
                        what = "polygons") %>%
  st_as_sf()

smb2019_pts <- smb2019 %>%
  st_as_sf(coords = c("lon1", "lat1"), crs = 4326)

grid_summary <- grid_sf %>%
  st_join(smb2019_pts %>% select(cod_n)) %>%
  group_by(x) %>%
  summarise(Cod = mean(cod_n, na.rm = TRUE), .groups = "drop")

ggplot() +
  geom_sf(data = grid_summary, aes(fill = Cod), colour = NA) +
  geom_sf(data = iceland_sf, fill = "grey60") +
  coord_sf(xlim = c(-27, -11), ylim = c(62.5, 67.5)) +
  scale_fill_viridis_c(option = "B", direction = -1,
                       name = "Mean cod (n)",
                       na.value = "white") +
  scale_x_continuous(NULL, NULL) +
  scale_y_continuous(NULL, NULL) +
  theme_bw()
```

---

::: {.panel .panel-warning}
::: panel-heading
Final exercise
:::
::: panel-body
Using the tools from this chapter:

1. Choose a species from `is_smb.csv` and a range of years.
2. Create a `facet_wrap()` map showing its spatial distribution for each year, using a gridded summary (`geom_tile()`) with an appropriate colour scale and transformation.
3. Add proper degree-labelled axes, a scale bar, and a north arrow.
4. Compose a final two-panel figure using `patchwork` that places your facetted map alongside a time series (line chart) of total annual biomass for the same species.
:::
:::
