---
title: "Mapping with ggplot2 — part 2"
---

# Setup

## Libraries

```{r setup}
library(tidyverse)
library(sf)
library(rnaturalearth)
library(rnaturalearthdata)
library(ggspatial)
library(patchwork)
```

## Data

```{r data}
minke_df <- read_csv("data/minke.csv")
smb      <- read_csv("data/smb_summary.csv")

# Convert minke whale positions to sf
minke_sf <- minke_df |>
  st_as_sf(coords = c("lon", "lat"), crs = 4326)

# Country polygons from rnaturalearth
iceland_sf <- ne_countries(country = "Iceland", scale = "medium")
world_sf   <- ne_countries(scale = "medium")

# Nephrops functional units — used for labelling examples
nfu <- read_sf("data/nephrops_fu.gpkg")

# Iceland coastline and depth contours — used in the mixed-CRS example
coast    <- read_sf("data/iceland_coastline.gpkg")
contours <- read_sf("data/iceland_contours.gpkg")

# SMB 2019 tows, as both a data frame and an sf object
smb2019     <- smb |> filter(year == 2019)
smb2019_pts <- smb2019 |>
  st_as_sf(coords = c("lon1", "lat1"), crs = 4326)
```

---

# From `geom_polygon()` to `geom_sf()`

In part 1 we used `map_data()` from the **maps** package to get country
outlines as plain data frames, plotted with `geom_polygon()`. That approach
works but has real limitations:

- it requires a `group` column to draw multi-part polygons correctly;
- it cannot represent lines or point geometries;
- the projection (`coord_quickmap()`) is only an approximation.

The modern approach is `geom_sf()`. It accepts any `sf` object — point, line,
or polygon — and pairs with `coord_sf()` for exact coordinate system handling.
Because the geometry is stored in the `sf` object itself, there is no need to
manage coordinate column names or polygon ordering manually.

```{r}
iceland_df <- map_data("world", region = "Iceland")

p_old <- ggplot() +
  geom_polygon(data = iceland_df,
               aes(long, lat, group = group),
               fill = "grey80", colour = "white") +
  coord_quickmap(xlim = c(-27, -11), ylim = c(62.5, 67.5)) +
  labs(title = "maps / geom_polygon (part 1 approach)") +
  theme_bw()

p_new <- ggplot() +
  geom_sf(data = iceland_sf, fill = "grey80", colour = "white") +
  coord_sf(xlim = c(-27, -11), ylim = c(62.5, 67.5)) +
  labs(title = "rnaturalearth / geom_sf (modern approach)") +
  theme_bw()

p_old + p_new
```

`geom_sf()` selects the correct rendering automatically (points, lines, or
filled polygons) based on the geometry type stored in the `sf` object. A
single call handles all geometry types.

---

# Basemaps with `rnaturalearth`

The **rnaturalearth** package provides country and coastline polygons as `sf`
objects at several resolutions (`"small"`, `"medium"`, `"large"`). It is a
well-maintained replacement for the **maps** package for basemap data.

```{r}
ggplot() +
  geom_sf(data = iceland_sf, fill = "grey80", colour = "white") +
  theme_bw()
```

Adding data on top follows the same layering logic as part 1: basemap first,
then data layers. Here we overlay the minke whale observations:

```{r}
ggplot() +
  geom_sf(data = iceland_sf, fill = "grey80") +
  geom_sf(data = minke_sf, aes(colour = sex), size = 2, alpha = 0.7) +
  scale_colour_brewer(name = "Sex", palette = "Set1") +
  scale_x_continuous(NULL, NULL) +
  scale_y_continuous(NULL, NULL) +
  theme_bw()
```

Zoom with `coord_sf(xlim, ylim)`. Values are in the units of the active CRS —
decimal degrees for the WGS84 default. Recall from part 1 that setting limits
inside `coord_sf()` zooms the *view* without clipping any features:

```{r}
ggplot() +
  geom_sf(data = world_sf, fill = "grey80") +
  geom_sf(data = minke_sf, aes(colour = sex), alpha = 0.8) +
  coord_sf(xlim = c(-24, -18), ylim = c(63, 65)) +
  scale_colour_brewer(name = "Sex", palette = "Set1") +
#  scale_x_continuous(NULL, NULL) +
#  scale_y_continuous(NULL, NULL) +
  theme_bw()
```

---

# Projections with `coord_sf()`

`coord_sf()` controls the CRS used when rendering all layers. Every layer is
automatically reprojected to match — you never need to reproject data manually
just for plotting. This connects directly to what you learned in the [CRS
chapter](crs.qmd): pass any EPSG code to the `crs` argument.

```{r fig.width=10, fig.height=4}
p_base <- ggplot() +
  geom_sf(data = iceland_sf, fill = "grey80") +
  geom_sf(data = minke_sf, aes(colour = sex), alpha = 0.7) +
  scale_colour_brewer(name = "Sex", palette = "Set1") +
  scale_x_continuous(NULL, NULL) +
  scale_y_continuous(NULL, NULL)

p1 <- p_base + coord_sf(crs = 4326) +
  labs(title = "WGS84 (geographic)", subtitle = "EPSG:4326")

p2 <- p_base + coord_sf(crs = 3057) +
  labs(title = "ISN93 / Lambert conformal", subtitle = "EPSG:3057")

p3 <- p_base + coord_sf(crs = st_crs("+proj=stere +lat_0=90 +lat_ts=65 +lon_0=-19")) +
  labs(title = "Polar stereographic", subtitle = "centred on Iceland")

p1 + p2 + p3
```

::: {.callout-note}
## Prefer EPSG codes over proj strings
The first two panels use EPSG codes (`crs = 4326`, `crs = 3057`). This is
more robust than raw proj strings (e.g. `"+proj=laea ..."`), which became
unreliable after PROJ 6. Use `crsuggest::suggest_crs()` to find the right
EPSG code for your area. Proj strings are still needed for custom projections
that have no EPSG entry, as in the third panel.
:::

## Mixing layers with different CRS

When layers come from different sources with different CRS, `coord_sf()`
aligns them automatically. Without an explicit `crs` argument it uses the CRS
of the first layer, which can produce unexpected results if the layer order
changes. Setting it explicitly is safer:

```{r}
# coast: Iceland coastline in a projected LAEA CRS
# contours: depth contours in WGS84 (EPSG:4326)
# — these would not align without coord_sf() handling the reprojection

# Without explicit CRS — display CRS taken from first layer (contours = WGS84)
p_implicit <- ggplot() +
  geom_sf(data = contours, colour = "steelblue", linewidth = 0.3) +
  geom_sf(data = coast, fill = "grey80") +
  theme_bw() +
  labs(title = "Implicit CRS (first layer wins)", subtitle = "WGS84")

# With explicit CRS — ISN2016 / LAEA Iceland (Iceland's official CRS)
p_explicit <- ggplot() +
  geom_sf(data = contours, colour = "steelblue", linewidth = 0.3) +
  geom_sf(data = coast, fill = "grey80") +
  coord_sf(crs = 8088) +
  theme_bw() +
  labs(title = "Explicit CRS", subtitle = "ISN2016 / LAEA Iceland (EPSG:8088)")

p_implicit + p_explicit
```

::: {.callout-important}
Always set `crs` explicitly in `coord_sf()` when combining layers from
different sources. The implicit default can change silently if you reorder
your layers.
:::

---

# Colour scales

Choosing an appropriate colour scale is one of the most consequential decisions
in map design. By this point in the course you are working with `geom_sf()` and
real polygon data, so the stakes are higher — a poor choice can completely hide
spatial structure.

## Continuous data: viridis palettes

The **viridis** palettes (`"viridis"`, `"magma"`, `"plasma"`, `"inferno"`,
`"cividis"`, `"mako"`, `"rocket"`, `"turbo"`) are built into ggplot2. They are
perceptually uniform, readable in greyscale, and colour-blind safe — a reliable
default for scientific maps:

```{r}
s <- smb |>
  filter(year == 2019) |>
  group_by(ir_lon, ir_lat) |>
  summarise(Cod = mean(cod_n), .groups = "drop") |>
  st_as_sf(coords = c("ir_lon", "ir_lat"), crs = 4326)

p_raster <- ggplot() +
  geom_sf(data = s, aes(colour = Cod), size = 3) +
  geom_sf(data = iceland_sf, fill = "grey60", colour = NA) +
  coord_sf(xlim = c(-27, -11), ylim = c(62.5, 67.5)) +
  scale_x_continuous(NULL, NULL) +
  scale_y_continuous(NULL, NULL) +
  theme_bw()

p1 <- p_raster + scale_colour_viridis_c(option = "D", name = "viridis")
p2 <- p_raster + scale_colour_viridis_c(option = "B", direction = -1,
                                         name = "magma (rev)")
p3 <- p_raster + scale_colour_viridis_c(option = "C", name = "plasma")

p1 + p2 + p3
```

## Colour scale transformations

Catch data are strongly right-skewed: most tows catch few fish and a few catch
a lot. On an untransformed scale, almost all the colour gradient is consumed by
the rare high values, obscuring variation among typical tows. Transforming the
scale spreads the gradient more evenly:

```{r}
p_log  <- p_raster +
  scale_colour_viridis_c(option = "B", direction = -1,
                         transform = "log1p", name = "log1p")
p_sqrt <- p_raster +
  scale_colour_viridis_c(option = "B", direction = -1,
                         transform = "sqrt", name = "sqrt")

p_log + p_sqrt
```

`log1p` (log(x + 1)) handles zeros and compresses high values strongly.
`sqrt` is a milder compression. Both typically reveal spatial structure that
a linear scale hides.

## Discrete intervals with `cut()`

Binning into labelled classes trades precision for legibility — often the
right choice for reports or presentations:

```{r}
s_cut <- s |>
  mutate(Cod_class = cut(Cod,
                         breaks = c(0, 23, 62, 145, Inf),
                         labels = c("Low", "Medium", "High", "Very high"),
                         include.lowest = TRUE))

ggplot() +
  geom_sf(data = s_cut, aes(colour = Cod_class), size = 3) +
  geom_sf(data = iceland_sf, fill = "grey60", colour = NA) +
  coord_sf(xlim = c(-27, -11), ylim = c(62.5, 67.5)) +
  scale_colour_viridis_d(option = "B", direction = -1,
                         name = "Cod abundance",
                         na.value = "grey80") +
  scale_x_continuous(NULL, NULL) +
  scale_y_continuous(NULL, NULL) +
  theme_bw() +
  labs(title = "Cod abundance classes — SMB 2019")
```

::: {.callout-note}
## Which colour scale to choose?

| Situation | Recommended approach |
|---|---|
| Precise values matter, data roughly symmetric | Continuous gradient |
| Data are strongly right-skewed (e.g. catch counts) | Continuous with `transform = "log1p"` or `"sqrt"` |
| Communicating to a non-technical audience or for a report | Discrete bins with `cut()` |
:::

---

# Labels: `geom_sf_text()` and `geom_sf_label()`

`geom_sf_text()` and `geom_sf_label()` place text at the geometry's position
(centroid for polygons). A reliable pattern is to compute centroids first with
`st_centroid()` and use those as a separate label layer:

```{r}
nfu_centroids <- st_centroid(nfu)

ggplot() +
  geom_sf(data = nfu, fill = NA, colour = "steelblue") +
  geom_sf_label(data = nfu_centroids,
                aes(label = name),
                size          = 2.5,
                label.padding = unit(0.15, "lines")) +
  theme_bw() +
  labs(title = "Nephrops functional units")
```

`geom_sf_text()` omits the label background box. Use it where you want
lighter annotation without the visual weight of a rectangle.

::: {.callout-tip}
## Exercise
Replace `geom_sf_label()` with `geom_sf_text()`. Then try mapping `name` to
`colour` to give each unit a distinct colour. Does it improve or clutter the
map?
:::

---

# Facetting

Facetting splits data into panels by one or more grouping variables. It is
particularly effective for showing how spatial distributions change over time.

## `facet_wrap()`

`facet_wrap()` arranges panels in a ribbon, filling rows left to right. Here
we convert the 2019 SMB tows to `sf` and map cod abundance for six survey
years:

```{r}
smb_sel_sf <- smb |>
  filter(year %in% c(1995, 2000, 2005, 2010, 2015, 2019)) |>
  st_as_sf(coords = c("lon1", "lat1"), crs = 4326)

ggplot() +
  geom_sf(data = iceland_sf, fill = "grey80") +
  geom_sf(data = smb_sel_sf, aes(size = cod_n),
          alpha = 0.15, colour = "red") +
  scale_size_area(max_size = 12, name = "Cod (n)") +
  coord_sf(xlim = c(-27, -11), ylim = c(62.5, 67.5)) +
  scale_x_continuous(NULL, NULL) +
  scale_y_continuous(NULL, NULL) +
  facet_wrap(~ year) +
  theme_bw()
```

## `facet_grid()`

`facet_grid(rows ~ cols)` creates a matrix of panels. This is useful for
comparing two variables simultaneously — here species and year. We pivot to
long format first so both species can act as a facet variable:

```{r}
smb_long_sf <- smb |>
  filter(year %in% c(2005, 2010, 2015, 2019)) |>
  pivot_longer(cols      = c(cod_n, haddock_n),
               names_to  = "species",
               values_to = "n") |>
  mutate(species = case_match(species,
                              "cod_n"     ~ "Cod",
                              "haddock_n" ~ "Haddock")) |>
  st_as_sf(coords = c("lon1", "lat1"), crs = 4326)

ggplot() +
  geom_sf(data = iceland_sf, fill = "grey80") +
  geom_sf(data = smb_long_sf, aes(size = n),
          alpha = 0.2, colour = "red") +
  scale_size_area(max_size = 8, name = "Catch (n)") +
  coord_sf(xlim = c(-27, -11), ylim = c(62.5, 67.5)) +
  scale_x_continuous(NULL, NULL) +
  scale_y_continuous(NULL, NULL) +
  facet_grid(species ~ year) +
  theme_bw()
```

In `facet_grid(rows ~ cols)`, the first variable defines rows and the second
defines columns. Use `. ~ year` to facet columns only, or `species ~ .` for
rows only. Note the `.` is a placeholder meaning "no variable".

::: {.callout-tip}
## Exercise
Add `saithe_n` as a third species to the `pivot_longer()` call. Does the
three-row grid remain legible, or would `facet_wrap()` be a better choice
here?
:::

---

# Axis formatting

## Geographic degree labels

`coord_sf()` draws graticule lines and labelled axes automatically. The
default labels are plain numbers; the helper functions below format them with
direction suffixes (N, S, E, W):

```{r}
scale_longitude <- function(min = -180, max = 180, step = 1, ...) {
  breaks <- seq(min, max, step)
  labels <- ifelse(breaks < 0,
                   paste0(abs(breaks), "\u00B0W"),
                   paste0(breaks,      "\u00B0E"))
  scale_x_continuous(name = NULL, breaks = breaks, labels = labels, ...)
}

scale_latitude <- function(min = -90, max = 90, step = 0.5, ...) {
  breaks <- seq(min, max, step)
  labels <- ifelse(breaks < 0,
                   paste0(abs(breaks), "\u00B0S"),
                   paste0(breaks,      "\u00B0N"))
  scale_y_continuous(name = NULL, breaks = breaks, labels = labels, ...)
}

ggplot() +
  geom_sf(data = iceland_sf, fill = "grey80") +
  coord_sf(xlim = c(-25, -12), ylim = c(63, 67)) +
  scale_longitude(min = -25, max = -12, step = 2) +
  scale_latitude( min =  63, max =  67, step = 1) +
  theme_bw() +
  labs(title = "Geographic degree labels")
```

## ICES statistical rectangle labels

ICES rectangles use a notation such as "40G1" rather than decimal degrees.
The functions below relabel axes accordingly. They require the **geo** package,
which is maintained by MFRI and available from GitHub:

```{r}
#| eval: false

# Install once:
# install.packages("remotes")
# remotes::install_github("hafro/geo")

library(mapplots)

d.ICES_rect <- expand.grid(
  # fixed reference points
  longitude = seq(-44 + 0.5, 68.5 - 0.5, 1),
  latitude = seq(36 + 0.25, 84.5 - 0.25, 0.5)
) |>
  mutate(ICES.rect = ices.rect2(
    d.ICES_rect$longitude, d.ICES_rect$latitude
  ))

f.plot_ICES <- function(
    xmin = -25, xmax = -12,
    ymin = 63, ymax = 67
      ){
  
  # Subset to map size
  d.ICES_rect_subset <- d.ICES_rect |>
    filter(
      between(longitude, xmin, xmax),
      between(latitude, ymin, ymax)
    ) |>
    mutate(
      long_ICES = stringr::str_sub(ICES.rect, start = 3, end = 4),
      lat_ICES = stringr::str_sub(ICES.rect, start = 1, end = 2)
    )
  
  # Axis labels
  lon_ICES <- unique(d.ICES_rect_subset$long_ICES)
  lat_ICES <- unique(d.ICES_rect_subset$lat_ICES)
  
  # Axis breaks
  lon_unique <- unique(d.ICES_rect_subset$longitude)
  lat_unique <- unique(d.ICES_rect_subset$latitude)
  
  # Gridding: vertical lines
  d.v_lines <- data.frame(
    longitude = seq(min(lon_unique) -0.5,
                    max(lon_unique) +0.5,
                    1) # every 1°!
    ) |>
    mutate(
      latitude_min = min(lat_unique) -0.25,
      latitude_max = max(lat_unique) +0.25
      )
  
  # Gridding: horizontal lines
  d.h_lines <- data.frame(
    latitude = seq(min(lat_unique) -0.25,
                   max(lat_unique) +0.25,
                    0.5) # every 0.5°!
  ) |>
    mutate(
      longitude_min = min(lon_unique) -0.5,
      longitude_max = max(lon_unique) +0.5
    )
  
  # Plot the plot
  ggplot() +
    geom_segment(
      data = d.v_lines,
      aes(x = longitude, xend = longitude, y = latitude_min, yend = latitude_max)
    ) +
    geom_segment(
      data = d.h_lines,
      aes(x = longitude_min, xend = longitude_max, y = latitude, yend = latitude)
    ) +
    geom_sf(data = iceland_sf, fill = "grey80") +
    coord_sf(xlim = c(xmin, xmax), ylim = c(ymin, ymax)) +
    scale_x_continuous(
      breaks = lon_unique,
      labels = lon_ICES
    ) +
    scale_y_continuous(
      breaks = lat_unique,
      labels = lat_ICES
    ) +
    theme_bw() +
    theme(
      axis.text.x = element_text(angle = 45, hjust = 1, vjust = 1),
      panel.grid.major = element_blank()
    )
  
  
}

f.plot_ICES()
f.plot_ICES(xmin = -40, xmax = -10, ymin = 50, ymax = 68)
```

---

# Map embellishments

## Scale bars and north arrows with `ggspatial`

The **ggspatial** package adds cartographic annotations to `geom_sf()` maps.
`annotation_scale()` draws a scale bar that updates automatically when you
zoom; `annotation_north_arrow()` adds a north arrow:

```{r}
ggplot() +
  geom_sf(data = iceland_sf, fill = "grey80") +
  annotation_scale(location = "bl", width_hint = 0.4) +
  annotation_north_arrow(
    location   = "bl",
    which_north = "true",
    pad_x      = unit(0.6, "in"),
    pad_y      = unit(0.4, "in"),
    style      = north_arrow_fancy_orienteering
  ) +
  coord_sf() +
  theme_bw() +
  labs(title = "Scale bar and north arrow")
```

## Themes for maps

`theme_bw()` is a good general-purpose theme that keeps axis labels and grid
lines, which aid spatial orientation. `theme_void()` strips everything — axes,
grid lines, and background — producing a cleaner result for publication figures
that already carry a scale bar and north arrow as their own spatial reference:

```{r}
ggplot() +
  geom_sf(data = iceland_sf, fill = "grey85", colour = "white") +
  geom_sf(data = minke_sf, aes(colour = sex), size = 1.8, alpha = 0.8) +
  scale_colour_brewer(name = "Sex", palette = "Set1") +
  annotation_scale(location = "br") +
  annotation_north_arrow(
    location = "tl",
    style    = north_arrow_minimal,
    height   = unit(0.8, "cm"),
    width    = unit(0.8, "cm")
  ) +
  coord_sf() +
  theme_void() +
  theme(legend.position          = "bottom",
        plot.title               = element_text(face = "bold")) +
  labs(title   = "Minke whale sightings",
       caption = "Source: MFRI")
```

Use `theme_void()` when the map is the entire figure and spatial context comes
from its own annotations. Use `theme_bw()` when the map sits alongside other
plots or when axis labels help readers orient themselves.

## Inset maps with `inset_element()`

`inset_element()` from **patchwork** places one plot inside another — useful
for a small context map inside a detailed regional map. Here we derive the
bounding box of the detail map programmatically so the red rectangle always
stays in sync:

```{r}
# Define the extent of the detail map explicitly so we can reuse it
detail_xlim <- c(-27, -11)
detail_ylim <- c(62.5, 67.5)

context_map <- ggplot() +
  geom_sf(data = world_sf, fill = "grey80", colour = "white",
          linewidth = 0.2) +
  geom_rect(aes(xmin = detail_xlim[1], xmax = detail_xlim[2],
                ymin = detail_ylim[1], ymax = detail_ylim[2]),
            fill = NA, colour = "red", linewidth = 1) +
  coord_sf(xlim = c(-40, 30), ylim = c(45, 80)) +
  theme_void()

detail_map <- ggplot() +
  geom_sf(data = iceland_sf, fill = "grey80") +
  geom_sf(data = minke_sf, aes(colour = sex), alpha = 0.8) +
  scale_colour_brewer(name = "Sex", palette = "Set1") +
  coord_sf(xlim = detail_xlim, ylim = detail_ylim) +
  theme_bw() +
  labs(title = "Minke whale sightings")

detail_map +
  inset_element(context_map,
                left   = 0.65, right  = 1.0,
                bottom = 0.0,  top    = 0.35)
```

::: {.callout-tip}
## Exercise
Move the inset to the top-left corner by adjusting the `left`, `right`,
`bottom`, and `top` arguments. Then try making the context map show a wider
extent (e.g. the full North Atlantic) and see whether the inset remains
informative at that scale.
:::

---

# Gridding point data with `st_make_grid()`

In part 1, the `grade()` function provided a quick way to bin points to a
regular lon–lat grid. A more rigorous approach uses `st_make_grid()` to
create proper `sf` polygon cells, then a spatial join to aggregate:

```{r}
bbox_sfc <- st_bbox(c(xmin = -27, xmax = -11,
                      ymin = 62.5, ymax = 67.5),
                    crs = 4326) |>
  st_as_sfc()

# Create the grid and label each cell with a unique ID
grid_sf <- st_make_grid(bbox_sfc,
                        cellsize = c(0.5, 0.25),
                        what     = "polygons") |>
  st_as_sf() |>
  mutate(cell_id = row_number())

# Spatial join: each tow point picks up the cell it falls in
grid_summary <- grid_sf |>
  st_join(smb2019_pts |> select(cod_n)) |>
  group_by(cell_id) |>                          # group by cell, not geometry
  summarise(Cod = mean(cod_n, na.rm = TRUE), .groups = "drop")

ggplot() +
  geom_sf(data = grid_summary, aes(fill = Cod), colour = NA) +
  geom_sf(data = iceland_sf, fill = "grey60") +
  coord_sf(xlim = c(-27, -11), ylim = c(62.5, 67.5)) +
  scale_fill_viridis_c(option    = "B",
                       direction = -1,
                       transform = "log1p",
                       name      = "Mean cod (n)\n(log scale)",
                       na.value  = "white") +
  scale_x_continuous(NULL, NULL) +
  scale_y_continuous(NULL, NULL) +
  theme_bw() +
  labs(title   = "Cod catch — spatial polygon grid (SMB 2019)",
       caption = "Grid: 0.5° × 0.25°; cells with no tows shown in white")
```

This approach is more rigorous than `grade()` because the grid cells are true
`sf` polygons with a known CRS, so the result integrates cleanly with other
spatial operations (e.g. clipping to an EEZ, spatial filtering).

::: {.callout-note}
## `grade()` vs `st_make_grid()`

| | `grade()` + `geom_tile()` | `st_make_grid()` + `st_join()` |
|---|---|---|
| Requires `sf`? | No | Yes |
| Result is spatial? | No (data frame) | Yes (sf object) |
| CRS-aware? | No | Yes |
| Suits further spatial ops? | No | Yes |
| Speed | Fast | Slower for large data |

Use `grade()` for quick exploration; use `st_make_grid()` when the gridded
output needs to participate in further spatial analysis or be exported as a
spatial file.
:::

::: {.callout-tip}
## Exercise
Change the grid resolution to `cellsize = c(1.0, 0.5)` (ICES rectangle size).
Compare the result to the ICES rectangle summary from part 1 — do they agree?
:::

---

# Case study: fishing intensity with `patchwork`

The OSPAR otter trawl intensity layer is a large, real-world polygon dataset
with strongly right-skewed values, making it an ideal test case for the
colour-scale techniques introduced above.

```{r ospar-data}
ospar_int <- read_sf("data/OSPAR_intensity_Otter_2015.gpkg") |>
  select(FishingH, KWfishingH, SurfSAR)
```

First, inspect the raw values to understand the skew:

```{r ospar-raw}
ggplot(ospar_int) +
  geom_sf(aes(fill = SurfSAR), colour = NA) +
  scale_fill_viridis_c(option = "plasma") +
  theme_bw() +
  labs(title    = "SurfSAR — raw scale",
       subtitle = "High skew obscures most spatial variation")
```

A log1p transform reveals the full spatial structure:

```{r ospar-log}
ggplot(ospar_int) +
  geom_sf(aes(fill = SurfSAR), colour = NA) +
  scale_fill_viridis_c(option    = "plasma",
                       direction = -1,
                       transform = "log1p",
                       name      = "SurfSAR (log)") +
  labs(title    = "Surface swept area ratio",
       subtitle = "OSPAR VMS data 2015",
       caption  = "Source: OSPAR Commission") +
  theme_bw()
```

## Comparing two effort metrics

`SurfSAR` is normalised by cell area; `KWfishingH` weights fishing hours by
vessel engine power. Placing them side by side with `patchwork` lets you
compare their spatial patterns directly:

```{r ospar-patchwork, fig.width=11, fig.height=5}
p_sar <- ggplot(ospar_int) +
  geom_sf(aes(fill = SurfSAR), colour = NA) +
  scale_fill_viridis_c(option    = "viridis",
                       direction = -1,
                       transform = "log1p",
                       name      = "SurfSAR (log)") +
  labs(title = "Surface swept area ratio") +
  theme_bw() +
  theme(axis.text  = element_blank(),
        axis.ticks = element_blank())

p_kw <- ggplot(ospar_int) +
  geom_sf(aes(fill = KWfishingH), colour = NA) +
  scale_fill_viridis_c(option    = "viridis",
                       direction = -1,
                       transform = "log1p",
                       name      = "kW·h (log)") +
  labs(title = "Effort (kW-fishing hours)") +
  theme_bw() +
  theme(axis.text  = element_blank(),
        axis.ticks = element_blank())

p_sar + p_kw +
  plot_annotation(
    title   = "Otter trawl fishing intensity — North-East Atlantic, 2015",
    caption = "Source: OSPAR Commission",
    theme   = theme(plot.title = element_text(face = "bold"))
  )
```

Both panels use the same viridis palette and direction so the colours are
directly comparable. Areas with many tows by small vessels can show high
`FishingH` but modest `KWfishingH`; a few large industrial vessels can
dominate `KWfishingH` while contributing fewer hours.

::: {.callout-note}
## Exercise

Work through these steps in order:

1. **Subset.** Filter the OSPAR layer to the North Sea (roughly −3° to 9°E,
   51° to 57°N) using `st_crop()`. Recreate the two-panel patchwork for this
   subregion. Does the spatial pattern look different at this finer scale?

2. **Extend.** The `select()` call in the data-loading chunk above drops
   `FishingH`. Add it back, then add a third panel showing `FishingH` on the
   same log1p / viridis scale. Use `plot_layout(ncol = 3)` to arrange all
   three panels in a single row.

3. **Reflect.** The two original panels used different viridis options
   (`plasma` vs `inferno`) in the original file. Both now use `"viridis"`.
   What was the problem with using different palettes for a side-by-side
   comparison, and when might deliberately different palettes be appropriate?
:::

---

# Final exercise

Bring together the techniques from both sessions:

1. Choose a species from `is_smb_biological.csv`. Join it to
   `is_smb_stations.csv` on `id`. Fill zeros for stations where the species
   was absent.
2. Convert the stations to `sf` and create a `facet_wrap()` map showing the
   species' spatial distribution for five or more years, using an
   `st_make_grid()` summary with an appropriate colour scale and
   transformation.
3. Add properly formatted degree-labelled axes (using `scale_longitude()` /
   `scale_latitude()`), a scale bar, and a north arrow.
4. Compose a final two-panel figure with `patchwork`: your facetted map on the
   left, and a time-series line chart of total annual catch for the same
   species on the right.
