---
title: "Mapping with ggplot2 — part 2"

---

# Getting started

## Libraries

```{r setup}
library(tidyverse)
library(sf)
library(rnaturalearth)
library(rnaturalearthdata)
library(ggspatial)
library(patchwork)
```

## Data

```{r data}
minke <- read_csv("data/minke.csv")
smb   <- read_csv("data/smb_summary.csv")

# Convert minke whale positions to sf
minke_sf <- minke |>
  st_as_sf(coords = c("lon", "lat"), crs = 4326)

# Country polygons from rnaturalearth
iceland_sf <- ne_countries(country = "Iceland", scale = "medium",
                           returnclass = "sf")
world_sf   <- ne_countries(scale = "medium", returnclass = "sf")

# Nephrops functional units — used for labelling examples
nfu <- read_sf("data/nephrops_fu.gpkg")

# Iceland coastline and depth contours — used in the mixed-CRS example
coast    <- read_sf("data/iceland_coastline.gpkg")
contours <- read_sf("data/iceland_contours.gpkg")

# SMB 2019 tows
smb2019 <- smb |> filter(year == 2019)
```

---

# From `ggplot1` to `geom_sf()`

In `ggplot1` we used `map_data()` from the **maps** package to get country
outlines as plain data frames, and plotted them with `geom_polygon()`. That
approach works, but it has limitations: it requires a specific data structure
with a column named `group` to draw polygons correctly, it cannot represent
lines or arbitrary polygon boundaries, and the projection (`coord_quickmap`) is
only an approximation.

The modern approach is `geom_sf()`. It accepts any sf object — point, line, or
polygon — and uses `coord_sf()` to manage the coordinate system exactly. Because
the geometry is stored in the sf object itself, there is no need to manage
coordinate column names or polygon ordering manually.

```{r}
# ggplot1 approach — a data frame with long/lat columns
iceland_df <- map_data("world", region = "Iceland")

p_old <- ggplot() +
  geom_polygon(data = iceland_df, aes(long, lat, group = group),
               fill = "grey80", colour = "white") +
  coord_quickmap(xlim = c(-27, -11), ylim = c(62.5, 67.5)) +
  labs(title = "maps / geom_polygon") +
  theme_bw()

# Modern approach — an sf object
p_new <- ggplot() +
  geom_sf(data = iceland_sf, fill = "grey80", colour = "white") +
  coord_sf(xlim = c(-27, -11), ylim = c(62.5, 67.5)) +
  labs(title = "rnaturalearth / geom_sf") +
  theme_bw()

p_old + p_new
```

`geom_sf()` selects the appropriate rendering (points, lines, or polygons)
automatically based on the geometry column. A single call handles all geometry
types — no need to choose between `geom_point()`, `geom_path()`, and
`geom_polygon()` depending on what you have.

---

# Basemaps with `rnaturalearth`

The **rnaturalearth** package provides country and coastline polygons as sf
objects at several resolutions (`"small"`, `"medium"`, `"large"`), and is a
well-maintained replacement for the **maps** package for basemap data.

```{r}
ggplot() +
  geom_sf(data = iceland_sf, fill = "grey80", colour = "white") +
  theme_bw()
```

Adding data on top follows exactly the same layering logic as `ggplot1`: the
basemap goes first, then data layers. Here we add the minke whale observations:

```{r}
ggplot() +
  geom_sf(data = iceland_sf, fill = "grey80") +
  geom_sf(data = minke_sf, aes(colour = sex), size = 2, alpha = 0.7) +
  scale_colour_brewer(name = "Sex", palette = "Set1") +
  labs(x = NULL, y = NULL) +
  theme_bw()
```

For a regional view, zoom with `coord_sf(xlim, ylim)`. Values are in the units
of the active CRS — decimal degrees for the WGS84 default:

```{r}
ggplot() +
  geom_sf(data = world_sf, fill = "grey80") +
  geom_sf(data = minke_sf, aes(colour = sex), alpha = 0.8) +
  coord_sf(xlim = c(-26, -12), ylim = c(62.5, 67.5)) +
  scale_colour_brewer(name = "Sex", palette = "Set1") +
  labs(x = NULL, y = NULL) +
  theme_bw()
```

---

# Projections with `coord_sf()`

`coord_sf()` controls the CRS used when rendering all map layers. Every layer is
automatically reprojected to match — you never need to reproject your data
manually just for plotting.

This connects directly to what you learned in the [CRS chapter](crs.qmd): pass
any EPSG code or proj string to the `crs` argument.

```{r}
p_base <- ggplot() +
  geom_sf(data = iceland_sf, fill = "grey80") +
  geom_sf(data = minke_sf, aes(colour = sex), alpha = 0.7) +
  scale_colour_brewer(name = "Sex", palette = "Set1") +
  scale_x_continuous(NULL, NULL) +
  scale_y_continuous(NULL, NULL)

p1 <- p_base + coord_sf(crs = 4326) +
  labs(title = "WGS84 (geographic)")

p2 <- p_base + coord_sf(crs = "+proj=laea +lat_0=65 +lon_0=-19") +
  labs(title = "LAEA centred on Iceland")

p3 <- p_base + coord_sf(crs = "+proj=stere +lat_0=90 +lat_ts=65 +lon_0=-19") +
  labs(title = "Stereographic")

p1 + p2 + p3
```

## Mixing layers with different CRS

When your layers have different CRS — a common situation when combining data
from different sources — `coord_sf()` aligns them automatically. The display CRS
defaults to the first layer's CRS, but setting it explicitly is best practice:

```{r}
# coast and contours are loaded in the setup chunk above
# coast: projected LAEA — contours: WGS84

# Explicit CRS avoids surprises when layer order changes
ggplot() +
  geom_sf(data = contours, colour = "steelblue", linewidth = 0.3) +
  geom_sf(data = coast,    fill = "grey80") +
  coord_sf(crs = 8088) +   # ISN2016 / LAEA Iceland — Iceland's official CRS
  theme_bw()
```

---

# Labels: `geom_sf_text()` and `geom_sf_label()`

For labelling polygon or point features, `geom_sf_text()` and `geom_sf_label()`
place text at the geometry's position (centroid for polygons). A useful pattern
is to compute centroid positions first with `st_centroid()` and use those as the
label layer:

```{r}
nfu_centroids <- st_centroid(nfu)

ggplot() +
  geom_sf(data = nfu, fill = NA, colour = "steelblue") +
  geom_sf_label(data = nfu_centroids, aes(label = name),
                size = 2.5, label.padding = unit(0.15, "lines")) +
  theme_bw()
```

`geom_sf_text()` omits the label box; use it for point labels where you do not
want the visual weight of a background rectangle.

---

# Facetting

Facetting splits data into panels by one or two grouping variables. It is
particularly effective for showing how species distributions shift over time.

## `facet_wrap()`

`facet_wrap()` arranges panels in a single ribbon, filling rows left to right.
Here, cod abundance is mapped for six survey years:

```{r}
smb_sel <- smb |> filter(year %in% c(1995, 2000, 2005, 2010, 2015, 2019))

ggplot() +
  geom_sf(data = iceland_sf, fill = "grey80") +
  geom_point(data = smb_sel, aes(lon1, lat1, size = cod_n),
             alpha = 0.15, colour = "red") +
  scale_size_area(max_size = 12, name = "Cod (n)") +
  coord_sf(xlim = c(-27, -11), ylim = c(62.5, 67.5)) +
  scale_x_continuous(NULL, NULL) +
  scale_y_continuous(NULL, NULL) +
  facet_wrap(~ year) +
  theme_bw()
```

## `facet_grid()`

`facet_grid(rows ~ cols)` creates a matrix of panels defined by two variables.
This is useful for comparing species and years simultaneously. We pivot the SMB
data to long format first so both species can be a facet variable:

```{r}
smb_long <- smb |>
  filter(year %in% c(2005, 2010, 2015, 2019)) |>
  pivot_longer(cols = c(cod_n, haddock_n),
               names_to  = "species",
               values_to = "n") |>
  mutate(species = recode(species, cod_n = "Cod", haddock_n = "Haddock"))

ggplot() +
  geom_sf(data = iceland_sf, fill = "grey80") +
  geom_point(data = smb_long, aes(lon1, lat1, size = n),
             alpha = 0.2, colour = "red") +
  scale_size_area(max_size = 8, name = "Catch (n)") +
  coord_sf(xlim = c(-27, -11), ylim = c(62.5, 67.5)) +
  scale_x_continuous(NULL, NULL) +
  scale_y_continuous(NULL, NULL) +
  facet_grid(species ~ year) +
  theme_bw()
```

In `facet_grid(rows ~ cols)`, the first variable defines rows and the second
defines columns. Use `. ~ year` for columns only, or `species ~ .` for rows only.

---

# Axis formatting

## Geographic degree labels

The helper functions below produce properly formatted degree–direction axis labels:

```{r}
scale_longitude <- function(min = -180, max = 180, step = 1, ...) {
  breaks <- seq(min, max, step)
  labels <- ifelse(breaks < 0,
                   paste0(abs(breaks), "\u00B0W"),
                   paste0(breaks, "\u00B0E"))
  scale_x_continuous(name = NULL, breaks = breaks, labels = labels, ...)
}

scale_latitude <- function(min = -90, max = 90, step = 0.5, ...) {
  breaks <- seq(min, max, step)
  labels <- ifelse(breaks < 0,
                   paste0(abs(breaks), "\u00B0S"),
                   paste0(breaks, "\u00B0N"))
  scale_y_continuous(name = NULL, breaks = breaks, labels = labels, ...)
}

ggplot() +
  geom_sf(data = iceland_sf, fill = "grey80") +
  coord_sf(xlim = c(-25, -12), ylim = c(63, 67)) +
  scale_longitude(min = -25, max = -12, step = 2) +
  scale_latitude(min = 63, max = 67, step = 1) +
  theme_bw()
```

## ICES statistical rectangle labels

ICES rectangles use a notation like "40G1" rather than decimal degrees. The functions below relabel axes accordingly. They require the **geo** package (`remotes::install_github("hafro/geo")`):

```{r eval=FALSE}
library(geo)

scale_longitude_ices <- function(min = -44, max = 68.5, step = 1, ...) {
  breaks <- seq(min + 0.5, max - 0.5, step)
  labels <- geo::d2ir(60, breaks) %>% stringr::str_sub(3)
  scale_x_continuous(name = NULL, breaks = breaks, labels = labels, ...)
}

scale_latitude_ices <- function(min = 36, max = 84.5, step = 0.5, ...) {
  breaks <- seq(min + 0.25, max - 0.25, step)
  labels <- geo::d2ir(breaks, 0) %>% stringr::str_sub(1, 2)
  scale_y_continuous(name = NULL, breaks = breaks, labels = labels, ...)
}

ggplot() +
  geom_sf(data = iceland_sf, fill = "grey80") +
  coord_sf(xlim = c(-25, -12), ylim = c(63, 67)) +
  scale_longitude_ices() +
  scale_latitude_ices() +
  theme(panel.grid.major = element_blank(),
        panel.grid.minor = element_line(linewidth = 0.5),
        axis.ticks = element_blank()) +
  theme_bw()
```

---

# Map embellishments

## Scale bars and north arrows with `ggspatial`

```{r}
ggplot() +
  geom_sf(data = iceland_sf, fill = "grey80") +
  annotation_scale(location = "bl", width_hint = 0.4) +
  annotation_north_arrow(
    location = "bl", which_north = "true",
    pad_x = unit(0.6, "in"), pad_y = unit(0.4, "in"),
    style = north_arrow_fancy_orienteering
  ) +
  coord_sf() +
  theme_bw()
```

## Themes for maps

`theme_void()` removes axes, grid lines, and background — often the cleanest choice for publication maps that carry their own spatial reference:

```{r}
ggplot() +
  geom_sf(data = iceland_sf, fill = "grey85", colour = "white") +
  geom_sf(data = minke_sf, aes(colour = sex), size = 1.8, alpha = 0.8) +
  scale_colour_brewer(name = "Sex", palette = "Set1") +
  annotation_scale(location = "br") +
  annotation_north_arrow(
    location = "tl",
    style = north_arrow_minimal,
    height = unit(0.8, "cm"), width = unit(0.8, "cm")
  ) +
  coord_sf() +
  theme_void() +
  theme(legend.position = "bottom",
        plot.title = element_text(face = "bold")) +
  labs(title = "Minke whale sightings", caption = "Source: MFRI")
```

## Inset maps with `inset_element()`

`inset_element()` places one plot inside another — useful for a small context map within a detailed regional map:

```{r}
context_map <- ggplot() +
  geom_sf(data = world_sf, fill = "grey80", colour = "white",
          linewidth = 0.2) +
  geom_rect(aes(xmin = -27, xmax = -11, ymin = 62.5, ymax = 67.5),
            fill = NA, colour = "red", linewidth = 1) +
  coord_sf(xlim = c(-40, 30), ylim = c(45, 80)) +
  theme_void()

detail_map <- ggplot() +
  geom_sf(data = iceland_sf, fill = "grey80") +
  geom_sf(data = minke_sf, aes(colour = sex), alpha = 0.8) +
  scale_colour_brewer(name = "Sex", palette = "Set1") +
  coord_sf() + theme_bw() +
  labs(title = "Minke whale sightings")

detail_map +
  inset_element(context_map,
                left = 0.65, right = 1.0,
                bottom = 0.0, top = 0.35)
```

---

# Gridding point data

## Using `grade()` for custom grids

The `grade()` function bins coordinates to a regular grid and returns the midpoint of each cell — useful for summarising survey data at a chosen resolution:

```{r}
grade <- function(x, dx) {
  if (dx > 1) warning("Not tested for grids larger than one degree")
  brks <- seq(floor(min(x)), ceiling(max(x)), dx)
  ints <- findInterval(x, brks, all.inside = TRUE)
  (brks[ints] + brks[ints + 1]) / 2
}

smb2019_gridded <- smb2019 |>
  mutate(glon = grade(lon1, 0.5),
         glat = grade(lat1, 0.25)) |>
  group_by(glon, glat) |>
  summarise(Cod = mean(cod_n), .groups = "drop")

ggplot() +
  geom_tile(data = smb2019_gridded,
            aes(x = glon, y = glat, fill = Cod)) +
  geom_sf(data = iceland_sf, fill = "grey60") +
  coord_sf(xlim = c(-27, -11), ylim = c(62.5, 67.5)) +
  scale_fill_viridis_c(option = "B", direction = -1,
                       transform = "log1p",
                       name = "Mean cod (n)") +
  scale_x_continuous(NULL, NULL) +
  scale_y_continuous(NULL, NULL) +
  theme_bw()
```

## Using `st_make_grid()` for spatial polygon grids

For a more rigorous approach that produces proper `sf` polygon grids respecting the CRS, use `st_make_grid()` then join with a spatial join:

```{r}
bbox_sfc <- st_bbox(c(xmin = -27, xmax = -11,
                      ymin = 62.5, ymax = 67.5),
                    crs = 4326) |>
  st_as_sfc()

grid_sf <- st_make_grid(bbox_sfc,
                        cellsize = c(0.5, 0.25),
                        what = "polygons") |>
  st_as_sf()

smb2019_pts <- smb2019 |>
  st_as_sf(coords = c("lon1", "lat1"), crs = 4326)

grid_summary <- grid_sf |>
  st_join(smb2019_pts |> select(cod_n)) |>
  group_by(x) |>
  summarise(Cod = mean(cod_n, na.rm = TRUE), .groups = "drop")

ggplot() +
  geom_sf(data = grid_summary, aes(fill = Cod), colour = NA) +
  geom_sf(data = iceland_sf, fill = "grey60") +
  coord_sf(xlim = c(-27, -11), ylim = c(62.5, 67.5)) +
  scale_fill_viridis_c(option = "B", direction = -1,
                       name = "Mean cod (n)",
                       na.value = "white") +
  scale_x_continuous(NULL, NULL) +
  scale_y_continuous(NULL, NULL) +
  theme_bw()
```

---

# Mapping fishing intensity with `patchwork`

The OSPAR otter trawl intensity layer (`OSPAR_intensity_Otter_2015.gpkg`) is a
large, real-world polygon dataset that rewards careful colour-scale choices.
With 91,190 cells and strongly right-skewed distributions, it illustrates several
techniques introduced in this chapter: `geom_sf()` with a continuous fill,
viridis colour scales, log transforms, and `patchwork` for multi-panel figures.

```{r ospar-ggplot-data, message=FALSE}
ospar_int <- read_sf("data/OSPAR_intensity_Otter_2015.gpkg") |>
  select(FishingH, KWfishingH, SurfSAR)
```

A basic starting map — note the heavy skew in the raw values:

```{r ospar-ggplot-raw, message=FALSE}
ggplot(ospar_int) +
  geom_sf(aes(fill = SurfSAR), colour = NA) +
  scale_fill_viridis_c(option = "plasma") +
  theme_bw() +
  labs(title = "SurfSAR — raw scale (hard to read)")
```

A log1p transform reveals the spatial structure across the whole range:

```{r ospar-ggplot-log, message=FALSE}
ggplot(ospar_int) +
  geom_sf(aes(fill = SurfSAR), colour = NA) +
  scale_fill_viridis_c(option = "plasma", direction = -1,
                       transform = "log1p",
                       name = "SurfSAR\n(log scale)") +
  labs(title = "Surface swept area ratio",
       subtitle = "OSPAR VMS data 2015",
       caption  = "Source: OSPAR Commission") +
  theme_bw()
```

## Side-by-side comparison with `patchwork`

Two effort metrics measure slightly different things: `SurfSAR` is already
normalised by cell area, whereas `KWfishingH` weighs fishing hours by vessel
engine power. The `patchwork` `+` operator places them side by side so spatial
patterns can be compared directly.

```{r ospar-patchwork, message=FALSE, fig.width=11, fig.height=5}
p_sar <- ggplot(ospar_int) +
  geom_sf(aes(fill = SurfSAR), colour = NA) +
  scale_fill_viridis_c(option    = "plasma",
                       direction = -1,
                       transform = "log1p",
                       name      = "SurfSAR (log)") +
  labs(title = "Surface swept area ratio") +
  theme_bw() +
  theme(axis.text = element_blank(), axis.ticks = element_blank())

p_kw <- ggplot(ospar_int) +
  geom_sf(aes(fill = KWfishingH), colour = NA) +
  scale_fill_viridis_c(option    = "inferno",
                       direction = -1,
                       transform = "log1p",
                       name      = "kW·h (log)") +
  labs(title = "Effort (kW-fishing hours)") +
  theme_bw() +
  theme(axis.text = element_blank(), axis.ticks = element_blank())

p_sar + p_kw +
  plot_annotation(
    title   = "Otter trawl fishing intensity — North-East Atlantic, 2015",
    caption = "Source: OSPAR Commission",
    theme   = theme(plot.title = element_text(face = "bold"))
  )
```

The two panels do not always agree: areas with many hours of fishing by small
vessels can have high `FishingH` but modest `KWfishingH`, whereas a few large
industrial vessels can dominate `KWfishingH` while contributing fewer hours.

::: {.callout-note}
## Exercise

1. Filter the OSPAR layer to cells in the North Sea only (roughly −3° to 9°E,
   51° to 57°N using `st_crop()`). Recreate the two-panel patchwork for this
   subregion. Does the spatial pattern look different at this scale?
2. Add a third panel to the patchwork showing `totvalue` (total landed value)
   on the same log1p scale. Use `plot_layout(ncol = 3)` to arrange all three
   in a single row.
3. The colour scales currently use different viridis options (`plasma` vs
   `inferno`). Change both to `viridis` with `direction = -1` so the panels are
   directly comparable. Does this make the comparison easier or harder to
   interpret?
:::

---

::: {.callout-note}
## Final exercise

Using the tools from this chapter:

1. Choose a species from `is_smb_biological.csv`. Join it to `is_smb_stations.csv`
   on `id` and fill zeros for stations where the species was absent.
2. Create a `facet_wrap()` map showing its spatial distribution for selected years,
   using a gridded summary (`geom_tile()`) with an appropriate colour scale and
   transformation.
3. Add proper degree-labelled axes, a scale bar, and a north arrow.
4. Compose a final two-panel figure using `patchwork` that places your facetted
   map alongside a time series (line chart) of total annual catch for the same
   species.
:::
