---
title: "Reading and writing spatial data"
execute:
  enabled: false
---

Spatial data is stored in files and in spatial databases. In this course we will focus on reading and writing files, with a brief look at spatial databases.

There is a large array of file formats for spatial data, both vector and raster. To get an idea, we can do this:

```{r setup}
library(tidyverse)
library(sf)
library(terra)   # modern raster I/O (replaces raster + rgdal)
```

```{r}
drivers_vec <- st_drivers("vector") %>% select(long_name)
nrow(drivers_vec)

drivers_rst <- st_drivers("raster") %>% select(long_name)
nrow(drivers_rst)
```

# GDAL

What we just saw are the file formats that can be read with the GDAL library.

The Geospatial Data Abstraction Library ([GDAL](https://gdal.org/)) is an external library for reading and writing raster and vector geospatial data in more than 200 formats and databases.

- Full list of vector drivers: <https://gdal.org/drivers/vector/index.html>
- Full list of raster drivers: <https://gdal.org/drivers/raster/index.html>

GDAL is the workhorse for reading and writing spatial data, used by many programs including ArcGIS and QGIS.

In R, the **sf** package provides `st_read()` / `st_write()` (and the tidyverse-friendly wrappers `read_sf()` / `write_sf()`) for vector data, linking directly to GDAL. For raster data, the modern approach is the **terra** package, which also links directly to GDAL.

::: {.callout-note}
The **rgdal** package, which was the original bridge to GDAL in R, was **retired from CRAN in October 2023**. Do not use it in new code. Its functionality is fully covered by **sf** (for vectors) and **terra** (for rasters).
:::

# Inspecting a file before reading

Before reading a file it is good practice to inspect its contents. For vector data (including GeoPackages, which can hold multiple layers) we use `st_layers()`:

```{r}
# A GeoPackage can hold multiple layers — inspect before reading
st_layers("./data_raw/Iceland_shapes.gpkg")
```

This tells us the layer names, geometry types, feature counts, and CRS — useful before committing to a full read.

# Common vector file formats

## Shapefiles

One of the most commonly used vector formats is the shapefile, developed by ESRI. A lot of the vector data you will find online is still in this format. However, it has significant limitations — see <http://switchfromshapefile.org/> for a full list.

A shapefile is not a single file, but several files with the same name and different extensions in the same folder. The three mandatory files are:

- `.shp` — shape format (geometric entities)
- `.shx` — index of the geometric entities
- `.dbf` — attributes (data)

Other files usually include a `.prj` file with the coordinate system.

Shapefile is **not** an open format and has known limitations: field names are truncated to 10 characters, it cannot store more than 255 columns, file size is limited to 2 GB, and it cannot mix geometry types.

```{r}
# Download and unzip a shapefile archive
dir.create("./data_raw", showWarnings = FALSE)
target_file <- "./data_raw/Iceland_shapefiles.zip"

download.file(
  url     = "https://ftp.hafro.is/pub/data/shapes/Iceland_shapefiles.zip",
  destfile = target_file
)

unzip(target_file, overwrite = TRUE, exdir = "./data_raw/")
file.remove(target_file)
```

Examine the `./data_raw/Iceland_shapes/` folder. You will see two sets of files with the same name and different extensions — these represent the coastline and depth contours for Iceland.

To read shapefiles we use `read_sf()` (tidyverse-friendly defaults) or `st_read()`:

```{r}
# Point directly to the .shp file
coast <- read_sf("./data_raw/Iceland_shapes/Iceland_coast.shp")

# Or use the dsn/layer form (no .shp extension needed)
coast <- read_sf(dsn = "./data_raw/Iceland_shapes/", layer = "Iceland_coast")

glimpse(coast)
```

## OGC GeoPackages (.gpkg)

A GeoPackage is a modern, open-format replacement for the shapefile, developed by the Open Geospatial Consortium. It uses a **single file** (`.gpkg`) based on SQLite.

Advantages over shapefiles:
- Single file — no risk of losing a companion `.dbf` or `.prj`
- Can store multiple layers (both vector and raster) in one file
- No field-name length restriction
- Open standard, supported by QGIS, ArcGIS, GDAL, and sf

```{r}
# Reading a specific layer from a GeoPackage
coast <- read_sf("./data_raw/iceland_coastline.gpkg")

# If the GeoPackage has multiple layers, specify which one
depth <- read_sf("./data_raw/iceland_shapes.gpkg", layer = "depth_contours")
```

GeoPackage is the **recommended format** for sharing vector data in this course.

## GeoJSON (.geojson / .json)

GeoJSON is a widely used open standard for encoding vector data using JSON. It is the default format for many web mapping APIs, and files are **automatically rendered as maps on GitHub**.

```{r}
# Read GeoJSON directly from a URL — no download needed
eez <- read_sf("https://raw.githubusercontent.com/datasets/geo-countries/master/data/countries.geojson")

# Write GeoJSON
write_sf(coast, "./data_raw/coast.geojson")
```

Limitations: GeoJSON is text-based, so it is less efficient for large datasets. It always uses WGS84 (EPSG:4326) by convention.

## FlatGeobuf (.fgb)

FlatGeobuf is a modern, high-performance binary format for vector data. It is streamable and supports spatial indexing, making it very efficient for reading subsets of large files — including directly from URLs without downloading the whole file.

```{r}
# Write to FlatGeobuf
write_sf(coast, "./data_raw/coast.fgb")

# Read back
coast_fgb <- read_sf("./data_raw/coast.fgb")

# Read a spatial subset using a bounding box — very efficient
my_bbox <- st_bbox(c(xmin = -25, xmax = -13, ymin = 63, ymax = 67),
                   crs = 4326)
coast_subset <- read_sf("./data_raw/coast.fgb", wkt_filter = st_as_text(st_as_sfc(my_bbox)))
```

FlatGeobuf is an excellent choice when working with large vector datasets.

## CSV (comma-separated values)

CSV is not a spatial format, but it is commonly used to store point data with coordinate columns.

```{r}
# Read a CSV with coordinates and convert to sf
minke <- read_csv("ftp://ftp.hafro.is/pub/data/csv/minke.csv") %>%
  st_as_sf(coords = c("lon", "lat"), crs = 4326)

glimpse(minke)
```

For point data this works well, but storing more complex geometries (lines, polygons) in CSV is impractical — use GeoPackage or GeoJSON instead.

# Reading subsets with SQL queries

When working with large files, it is often more efficient to read only the rows you need. `st_read()` supports SQL queries via the `query` argument:

```{r}
# Read only surveys from Iceland from a large dataset
# (the table name is the layer name inside the file)
iceland_surveys <- st_read(
  "./data_raw/datras_surveys.gpkg",
  query = "SELECT * FROM datras_surveys WHERE country = 'Iceland'"
)

# Or filter spatially using a bounding box
coast_west <- st_read(
  "./data_raw/Iceland_shapes/Iceland_coast.shp",
  wkt_filter = "POLYGON ((-25 63, -18 63, -18 67, -25 67, -25 63))"
)
```

This avoids loading millions of features into memory when you only need a subset.

# Raster I/O with **terra**

The **terra** package is the modern replacement for **raster** (and **rgdal**) for raster data. It is faster, has a cleaner API, and links directly to GDAL.

## Reading rasters

```{r}
# Read a single-band GeoTIFF
mintemp <- rast("ftp://ftp.hafro.is/pub/data/rasters/Iceland_minbtemp.tif")

# Read a multi-band raster (RasterStack equivalent)
temps <- rast(c(
  "ftp://ftp.hafro.is/pub/data/rasters/Iceland_minbtemp.tif",
  "ftp://ftp.hafro.is/pub/data/rasters/Iceland_maxbtemp.tif"
))

print(mintemp)
plot(mintemp)
```

## Cloud Optimised GeoTIFF (COG)

A Cloud Optimised GeoTIFF (COG) is a standard GeoTIFF with an internal structure that allows efficient partial reads from cloud storage (e.g. AWS S3, Google Cloud Storage) without downloading the whole file. **terra** reads COGs transparently — just supply the URL.

```{r}
# Read directly from a remote COG — no download required
# (example URL; replace with a real COG endpoint)
cog <- rast("/vsicurl/https://example.org/data/iceland_dem.tif")

# Read only a spatial subset (very efficient with COGs)
my_ext <- ext(-25, -18, 63, 67)  # xmin, xmax, ymin, ymax
cog_crop <- crop(cog, my_ext)
```

The `/vsicurl/` prefix tells GDAL to read over HTTP. Similar prefixes exist for `/vsis3/` (AWS S3), `/vsigs/` (Google Cloud Storage), and `/vsiaz/` (Azure Blob Storage).

## Writing rasters

```{r}
# Write a raster to GeoTIFF (the most common format)
writeRaster(mintemp, "./data/mintemp_output.tif", overwrite = TRUE)

# Write as a Cloud Optimised GeoTIFF
writeRaster(mintemp, "./data/mintemp_cog.tif",
            filetype = "COG",
            overwrite = TRUE)

# Write a NetCDF (common in oceanography and climate science)
writeRaster(temps, "./data/temps.nc", overwrite = TRUE)
```

## Raster metadata without reading data

```{r}
# Inspect raster metadata without loading all the cell values
describe("ftp://ftp.hafro.is/pub/data/rasters/Iceland_minbtemp.tif")
```

# Writing vector data

To write vector data, use `write_sf()` or `st_write()`. Both can guess the driver from the file extension.

```{r}
# Write to GeoPackage (recommended)
write_sf(coast, "./data/coast.gpkg")

# Write to shapefile
write_sf(coast, "./data/coast.shp")

# Write to GeoJSON
write_sf(coast, "./data/coast.geojson")

# Overwrite an existing file without prompting
write_sf(coast, "./data/coast.gpkg", delete_dsn = TRUE)
```

## Writing point data to CSV

If you want to export an `sf` point object to CSV and **include the coordinates** as columns, pass the `GEOMETRY=AS_XY` layer option:

```{r}
mypts <- data.frame(lon = c(-11.77, -11.76, -11.744),
                    lat = c(66.02, 66.018, 66.014)) %>%
  st_as_sf(coords = c("lon", "lat"), crs = 4326)

dir.create("./data", showWarnings = FALSE)
write_sf(mypts, "./data/pts.csv", layer_options = "GEOMETRY=AS_XY")
```

Alternatively, use `st_coordinates()` to extract the coordinates as a data frame and bind them to the attributes before writing:

```{r}
mypts %>%
  bind_cols(st_coordinates(.) %>% as_tibble() %>% rename(lon = X, lat = Y)) %>%
  st_drop_geometry() %>%
  write_csv("./data/pts_tidy.csv")
```

# Spatial databases

Beyond flat files, spatial data is often stored in spatial databases. The two most common are:

## SQLite / SpatiaLite

SpatiaLite adds spatial extensions to SQLite. A GeoPackage is actually a specialised SpatiaLite database. `sf` can connect to SpatiaLite databases directly:

```{r}
# Connect to a SpatiaLite database and read a layer
conn <- DBI::dbConnect(RSQLite::SQLite(), "./data/my_spatialite.db")
layer <- read_sf(conn, "my_layer_name")
DBI::dbDisconnect(conn)
```

## PostGIS

PostGIS is a spatial extension to PostgreSQL and is the standard for server-based spatial databases. Reading from PostGIS with `sf` works via a DBI connection:

```{r eval=FALSE}
library(RPostgres)

conn <- DBI::dbConnect(
  RPostgres::Postgres(),
  dbname = "my_database",
  host   = "localhost",
  user   = "my_user",
  password = "my_password"
)

# Read a spatial table
survey_stations <- read_sf(conn, "survey_stations")

# Or use an SQL query for a subset
iceland_stations <- read_sf(
  conn,
  query = "SELECT * FROM survey_stations WHERE country = 'Iceland'"
)

DBI::dbDisconnect(conn)
```

# Quick reference: formats and when to use them

| Format | Extension | Type | Use when... |
|---|---|---|---|
| GeoPackage | `.gpkg` | Vector | General use; sharing data; multiple layers |
| Shapefile | `.shp` + others | Vector | Receiving legacy data; collaborating with ArcGIS users |
| GeoJSON | `.geojson` | Vector | Web mapping; sharing on GitHub; APIs |
| FlatGeobuf | `.fgb` | Vector | Large datasets; streaming; cloud access |
| GeoTIFF | `.tif` | Raster | General raster storage |
| Cloud Optimised GeoTIFF | `.tif` | Raster | Cloud/remote access; large rasters |
| NetCDF | `.nc` | Raster | Oceanographic / climate model output |
| CSV | `.csv` | Points only | Simple point data; exchange with non-GIS software |

---

::: {.panel .panel-warning}
::: panel-heading
Exercise 1 — Reading vector data
:::
::: panel-body
1. Download the Iceland shapefiles, read the coastline into R, and inspect its metadata using `st_crs()`, `st_bbox()`, and `st_geometry_type()`.
2. Write the coastline to a GeoPackage and a GeoJSON file.
3. Read the GeoJSON file back in and confirm it has the same features using `nrow()`.
:::
:::

::: {.panel .panel-warning}
::: panel-heading
Exercise 2 — Reading raster data
:::
::: panel-body
1. Using **terra**, read the Iceland minimum bottom temperature raster.
2. Check its resolution, extent, and CRS.
3. Crop it to a bounding box covering only the area west of -20°W and write the result as a new GeoTIFF.
:::
:::
