---
title: "Interactive maps"
execute:
  enabled: true
---

**Needed libraries**:

```{r setup}
library(tidyverse)
library(sf)
library(terra)
library(rnaturalearth)
library(leaflet)
library(leaflet.extras)
library(leafem)
library(leafpop)
library(mapview)
library(tmap)
library(ggiraph)
library(htmlwidgets)
```

R has a mature ecosystem for building interactive maps that can be embedded in HTML
reports, Quarto documents, Shiny applications, or saved as standalone web pages.
This chapter covers the main approaches, from quick exploratory tools to
high-performance solutions for large datasets.

A key technical point before starting: almost all interactive map packages render
in the browser using the **WGS84** geographic coordinate system (EPSG:4326 —
longitude/latitude). When your *sf* objects are in a projected CRS, most packages
will reproject them silently, but it is good practice to be explicit and to verify
that all layers share a common CRS before plotting.

---

# Quick exploration with `mapview`

The `mapview` package is the fastest way to get a spatial object on an interactive
map — it requires almost no setup and works directly with *sf* and *terra* objects.
It is ideal for data exploration during analysis, not necessarily for polished outputs.

**Data used**:
```{r}
minke <- read_csv("data/minke.csv")

minke_sf <- minke |>
  st_as_sf(coords = c("lon", "lat"), crs = 4326)
```

The simplest possible interactive map:
```{r}
mapview(minke_sf)
```

Use `zcol` to colour points by an attribute variable, and `layer.name` to label
the legend:
```{r}
mapview(minke_sf, zcol = "sex", layer.name = "Sex")
```

Multiple *sf* layers can be combined with `+`. The result includes a layer control
panel so the user can toggle each layer on and off:
```{r}
nfu <- read_sf("data/nephrops_fu.gpkg")

mapview(nfu, layer.name = "Nephrops FU", alpha.regions = 0.2) +
  mapview(minke_sf, zcol = "sex", layer.name = "Minke whales")
```

To view two maps side by side with synchronised pan and zoom, use `sync()`:
```{r}
#| eval: false
m1 <- mapview(minke_sf, zcol = "sex",     layer.name = "Sex")
m2 <- mapview(minke_sf, zcol = "age",     layer.name = "Age")
# sync(m1, m2)
m1 | m2
```

`mapview` also accepts `terra` raster objects directly:
```{r}
mintemp <- terra::rast("data/Iceland_minbtemp.tif")

mapview(mintemp, layer.name = "Min. bottom temp.")
```

::: {.callout-note}
## Exercise

Load the ICES ecoregions shapefile (`ices_ecoregions.gpkg`) and the Nephrops survey
data (`datras_2018_haul.csv`). Create a combined `mapview` showing the ecoregions
coloured by name, and the survey station locations on top.
:::

---

# The `tmap` package

`tmap` uses a grammar-of-graphics approach similar to ggplot2, and can produce
both static and interactive maps by switching a single mode setting. This makes it
easy to iterate between print-quality figures and interactive exploratory maps
using the same code.

```{r}
# Interactive mode — renders in the browser
tmap_mode("view")

# Switch back to static mode for reports/print
# tmap_mode("plot")
```

The basic building block is `tm_shape()`, which specifies the data, followed by
layer functions such as `tm_polygons()`, `tm_dots()`, `tm_lines()`, etc.:

```{r}
ices_er <- read_sf("data/ices_ecoregions.gpkg")

tm_shape(ices_er) +
  tm_polygons(col = "ecoregion",
              alpha = 0.5,
              title = "Ecoregion",
              id = "ecoregion")    # shown in the tooltip on hover
```

Adding multiple layers works the same way as in ggplot — each `tm_shape()` call
introduces a new dataset:

```{r}
tm_shape(ices_er) +
  tm_polygons(col = "ecoregion", fill_alpha = 0.4, id = "ecoregion") +
  tm_shape(minke_sf) +
  tm_dots(col = "sex",
          palette = c(Male = "steelblue", Female = "tomato"),
          size = 1,
          id = "sex",
          popup.vars = c("age", "length", "weight"))  # shown in click popup
```

To produce a static version of the same map, simply switch the mode:
```{r}
tmap_mode("plot")

tm_shape(ices_er) +
  tm_polygons(col = "ecoregion", alpha = 0.4) +
  tm_shape(minke_sf) +
  tm_dots(col = "sex",
          palette = c(Male = "steelblue", Female = "tomato"),
          size = 0.1) +
  tm_layout(legend.outside = TRUE)
```

```{r}
# Switch back for the rest of this chapter
tmap_mode("view")
```

::: {.callout-note}
## Exercise

Using `tmap`, create an interactive map of the Nephrops functional units
(`nephrops_fu.gpkg`). Colour the polygons by `name` and configure the tooltip to
show both `name` and the area of each unit. Then switch to `tmap_mode("plot")` and
reproduce the same map as a static figure suitable for a report.
:::

---

# `leaflet` — full control over interactive maps

For polished, production-quality interactive maps, the `leaflet` package gives the
most control. It is more verbose than `mapview` or `tmap`, but every visual element
can be customised precisely.

Leaflet works by piping layers together with `|>`. Each `add*()` call adds a new
layer to the map widget.

## The basic construct

```{r}
leaflet() |>
  addTiles() |>
  addPopups(lng = -21.9426,
            lat =  64.1355,
            popup = "Reykjavík")
```

Here:

- `leaflet()` creates the map widget
- `addTiles()` adds the default OpenStreetMap background
- `addPopups()` adds a labelled marker

## Data input

Leaflet accepts *sf* objects directly — you do not need to specify `lng` and `lat`
when the object already has a geometry column:

```{r}
leaflet() |>
  addTiles() |>
  addCircles(data = minke_sf, radius = 10000)
```

For plain data frames you must supply `lng` and `lat` explicitly:

```{r}
leaflet() |>
  addTiles() |>
  addCircles(data = minke,
             lng  = ~lon,
             lat  = ~lat,
             radius = 10000)
```

Note the `~` prefix: in leaflet, all references to column names in the data must
be written as one-sided formulas (e.g. `~lon`, `~sex`). This is different from
ggplot, where you use `aes()`.

## Layer functions and their ggplot equivalents

```
leaflet            ggplot2
---------          -------
addCircles()       geom_point()
addCircleMarkers() geom_point()   (fixed pixel size, not metres)
addPolylines()     geom_path()
addPolygons()      geom_polygon()
addRasterImage()   geom_raster()
```

Key aesthetic arguments:

```
leaflet                    ggplot2
---------                  -------
lng / lat                  x / y
color                      colour
fillColor                  fill
weight                     linewidth
radius                     size
opacity / fillOpacity      alpha
label                      (hover tooltip)
popup                      (click popup — supports HTML)
```

## Mapping variables to colours

Leaflet does not automatically assign colours to categorical variables the way
ggplot does. You must first create a colour palette function using one of the
`color*()` helpers, then pass it to the `color` argument.

| Helper | Use case |
|---|---|
| `colorFactor()` | Categorical (factor) data |
| `colorNumeric()` | Continuous data, linear mapping |
| `colorBin()` | Continuous data, binned into intervals |
| `colorQuantile()` | Continuous data, quantile-based bins |

```{r}
pal_sex <- colorFactor(palette = c("steelblue", "tomato"),
                       domain   = c("Male", "Female"))

leaflet(minke_sf) |>
  addTiles() |>
  addCircles(color       = ~pal_sex(sex),
             radius      = 8000,
             fillOpacity = 0.8,
             weight      = 1,
             label       = ~sex) |>
  addLegend(position = "bottomleft",
            pal      = pal_sex,
            values   = ~sex,
            title    = "Sex")
```

For the heavily skewed abundance data common in fisheries surveys, `colorBin()` with
log-transformed breaks is often more informative than a linear mapping:

```{r}
smb2019 <- read_csv("data/smb_summary.csv") |> 
  filter(year == 2019) |>
  st_as_sf(coords = c("lon1", "lat1"), crs = 4326)

pal_cod <- colorBin(palette = "YlOrRd",
                    domain  = smb2019$cod_n,
                    bins    = c(0, 10, 50, 200, 500, 3000),
                    na.color = "transparent")

leaflet(smb2019) |>
  addProviderTiles(providers$Esri.OceanBasemap) |>
  addCircleMarkers(radius      = 6,
                   color       = "white",
                   weight      = 0.5,
                   fillColor   = ~pal_cod(cod_n),
                   fillOpacity = 0.85,
                   label       = ~as.character(round(cod_n)),
                   popup       = ~paste0("<b>Station:</b> ", tow_nr,
                                         "<br><b>Cod (n):</b> ", cod_n)) |>
  addLegend(position = "bottomright",
            pal      = pal_cod,
            values   = ~cod_n,
            title    = "Cod abundance (n)")
```

Note the `popup` argument: unlike `label` (which appears on hover), a popup
appears on click and supports arbitrary HTML markup.

::: {.callout-note}
## Exercise

Modify the map above so that:

- The circle radius is proportional to `cod_n` (not fixed at 6).
- The popup also shows `haddock_n` alongside `cod_n`.
- Use `colorQuantile()` instead of `colorBin()` and compare the two maps.
  Which one communicates the spatial pattern more clearly, and why?
:::

## Basemaps

The default OpenStreetMap tiles (`addTiles()`) are not always suitable for marine
or fisheries data. `addProviderTiles()` gives access to a large library of
alternative basemaps. A full searchable preview is at
<https://leaflet-extras.github.io/leaflet-providers/preview/>.

```{r}
m_base <- leaflet() |>
  setView(lng = -19, lat = 65, zoom = 6)

# Ocean-focused basemaps
m_base |> addProviderTiles(providers$Esri.OceanBasemap)

m_base |> addProviderTiles(providers$Esri.WorldImagery)

m_base |> addProviderTiles(providers$CartoDB.Positron)
```

Often no basemap at all gives the cleanest result for oceanographic data, because
a detailed land/ocean basemap can distract from the data itself. In that case, load
a coastline polygon as a plain `addPolygons()` layer instead:

```{r}
iceland_sf <- ne_countries(country = "Iceland", scale = 10, returnclass = "sf")

leaflet() |>
  addPolygons(data   = iceland_sf,
              color  = "grey40",
              weight = 1,
              fillColor = "grey80",
              fillOpacity = 1) |>
  addCircles(data        = minke_sf,
             color       = ~pal_sex(sex),
             radius      = 8000,
             fillOpacity = 0.8,
             weight      = 0,
             label       = ~sex) |>
  addLegend("bottomleft", pal = pal_sex, values = minke_sf$sex,
            title = "Sex")
```

## Layer control

When displaying multiple layers, `addLayersControl()` adds a toggle panel that
allows the user to show or hide individual layers. Assign each layer a `group`
name and pass those names to either `baseGroups` (radio button — only one active
at a time) or `overlayGroups` (checkboxes — any combination):

```{r}
# NOTE: BAD EXAMPLE
trail <- read_csv("data/smb_summary.csv") |>
  filter(year %in% 2001:2009) |> 
  st_as_sf(coords = c("lon1", "lat1"), crs = 4326) |>
  group_by(year) |>
  summarise(do_union = FALSE) |>
  st_cast("LINESTRING") |> 
  mutate(year = as.character(year))

pal_vessel <- colorFactor(
  palette = RColorBrewer::brewer.pal(n = length(unique(trail$year)), name = "Set1"),
  domain  = trail$year)

leaflet() |>
  addProviderTiles(providers$Esri.OceanBasemap, group = "Ocean") |>
  addProviderTiles(providers$Esri.WorldImagery,  group = "Satellite") |>
  addPolylines(data    = trail,
               color   = ~pal_vessel(year),
               weight  = 2,
               opacity = 1,
               label   = ~year,
               group   = ~year) |>
  #addLegend("bottomleft", pal = pal_vessel,
  #          values = trail$year, title = "Vessel") |>
  addLayersControl(
    baseGroups    = c("Ocean", "Satellite"),
    overlayGroups = unique(trail$year),
    options       = layersControlOptions(collapsed = FALSE))
```

## Polygons with rich popups

The `leafpop` package makes it easy to attach formatted tables or even ggplot
figures to polygon popups:

```{r}
nfu_proj <- nfu |> st_transform(4326)

# A simple summary table popup per polygon
leaflet(nfu_proj) |>
  addTiles() |>
  addPolygons(color       = "steelblue",
              weight      = 1,
              fillOpacity = 0.3,
              popup       = popupTable(nfu_proj,
                                       zcol = c("name"),
                                       row.numbers = FALSE,
                                       feature.id  = FALSE))
```

## Raster layers

Use `addRasterImage()` from the `leafem` package for *terra* `SpatRaster` objects.
For very large rasters, `leafem::addGeotiff()` streams tiles on demand and is
much faster than loading the full raster into memory:

```{r}
mintemp_r <- terra::rast("data/Iceland_minbtemp.tif")

pal_temp <- colorNumeric(palette   = "RdYlBu",
                         domain    = terra::values(mintemp_r),
                         na.color  = "transparent",
                         reverse   = TRUE)

leaflet() |>
  addProviderTiles(providers$Esri.OceanBasemap) |>
  #leafem::addRasterRGB(data = mintemp_r) |>      # quick RGB preview, or use addRasterImage:
  addRasterImage(mintemp_r,
                 colors  = pal_temp,
                 opacity = 0.8,
                 group   = "Min. bottom temp.") |>
  addLegend(pal    = pal_temp,
            values = terra::values(mintemp_r),
            title  = "Min. bottom temp. (°C)") |>
  addLayersControl(overlayGroups = "Min. bottom temp.")
```

## Point clustering

When displaying thousands of points, overlapping markers make the map unreadable.
`addMarkerClusterOptions()` from `leaflet.extras` groups nearby points into
clusters that expand on click:

```{r}
all_smb <- read_csv("data/is_smb_stations.csv") |>
  st_as_sf(coords = c("lon1", "lat1"), crs = 4326)

leaflet(all_smb) |>
  addProviderTiles(providers$CartoDB.Positron) |>
  addCircleMarkers(radius      = 4,
                   color       = "steelblue",
                   fillOpacity = 0.6,
                   weight      = 0,
                   label       = ~as.character(year),
                   clusterOptions = markerClusterOptions())
```

## Drawing tools and measurement

`leaflet.extras` also adds user-facing drawing tools and a measurement widget,
which are useful for field data collection workflows or for letting users define
an area of interest interactively:

```{r}
leaflet() |>
  addProviderTiles(providers$Esri.OceanBasemap) |>
  setView(lng = -19, lat = 65, zoom = 6) |>
  addDrawToolbar(
    targetGroup    = "drawn",
    editOptions    = editToolbarOptions(selectedPathOptions =
                                          selectedPathOptions())) |>
  addMeasure(
    position        = "bottomleft",
    primaryLengthUnit  = "kilometers",
    primaryAreaUnit    = "sqmeters") |>
  addLayersControl(overlayGroups = "drawn")
```

## Saving as a standalone HTML file

Any leaflet (or mapview / tmap in view mode) widget can be saved as a
self-contained HTML file. The `selfcontained = TRUE` argument embeds all
JavaScript and CSS so the file works offline:

```{r}
#| eval: false
m <- leaflet(minke_sf) |>
  addTiles() |>
  addCircles(color = ~pal_sex(sex), radius = 8000)

htmlwidgets::saveWidget(m, "minke_map.html", selfcontained = TRUE)
```

In a Quarto document, set `embed-resources: true` in the YAML header to achieve
the same effect for the whole rendered document.

::: {.callout-note}
## Exercise

Build a leaflet map of the DATRAS 2018 haul data (`datras_2018_haul.csv`) that:

1. Shows each tow as a line segment (start to end position).
2. Colours the lines by `survey` using a `colorFactor` palette.
3. Adds a popup on each line showing the survey name and the haul date.
4. Includes a basemap toggle between `Esri.OceanBasemap` and `CartoDB.Positron`.
5. Saves the result as a standalone HTML file.
:::

---

# `ggiraph` — adding interactivity to ggplot maps

If you are already comfortable with ggplot2, `ggiraph` is the easiest path to
interactivity. It works by replacing standard geoms with interactive equivalents
(`geom_sf_interactive()`, `geom_point_interactive()`, etc.) that accept `tooltip`
and `onclick` aesthetics. The `girafe()` function then renders the plot as an
interactive SVG.

```{r}
ices_er_int <- ices_er |>
  mutate(tooltip_text = paste0("<b>", ecoregion, "</b>",
                               "<br>Area: ", round(area_km2), " km²"))

p <- ggplot(ices_er_int) +
  geom_sf_interactive(aes(fill       = area_km2,
                           tooltip    = tooltip_text,
                           data_id    = ecoregion),
                      colour = "white", linewidth = 0.3) +
  scale_fill_viridis_c(name = "Area (km²)", trans = "log10") +
  theme_bw()

girafe(ggobj = p,
       options = list(
         opts_hover(css = "fill-opacity:0.7;stroke:black;stroke-width:1.5px;"),
         opts_tooltip(css = "background:white;padding:5px;border-radius:4px;")))
```

`ggiraph` is particularly useful when you need a publication-quality static map
that also works interactively in an HTML report, since the same ggplot code
produces both.

::: {.callout-note}
## Exercise

Using `ggiraph`, recreate the SMB 2019 cod bubble map from the ggplot chapter as
an interactive plot. The tooltip should show station ID, year, and cod abundance.
Hover highlighting should change the fill colour of the hovered point.
:::

---

# Performance with large datasets

Standard leaflet renders all geometries in the browser DOM, which becomes slow
above roughly 10,000 features. For larger datasets, consider these approaches:

**Point clustering** (shown above with `markerClusterOptions()`) is the simplest
fix for dense point data.

**`leafgl`** uses WebGL to render *sf* objects with hundreds of thousands of
features at interactive frame rates, with a nearly identical API to base leaflet:

```{r eval=FALSE}
library(leafgl)

# Render a large point dataset via WebGL
big_pts <- read_csv("data/is_smb_stations.csv") |>
  st_as_sf(coords = c("lon1", "lat1"), crs = 4326)

leaflet() |>
  addProviderTiles(providers$CartoDB.DarkMatter) |>
  addGlPoints(data       = big_pts,
              fillColor  = "steelblue",
              fillOpacity = 0.5,
              radius     = 6)
```

**`mapgl`** wraps MapLibre GL JS (the open-source fork of Mapbox GL JS, requiring
no API key) and can handle millions of features via vector tiles. It is the most
capable option for production applications:

```{r eval=FALSE}
library(mapgl)

maplibre() |>
  add_circle_layer(id     = "smb-stations",
                   source = big_pts,
                   circle_color  = "steelblue",
                   circle_radius = 4,
                   circle_opacity = 0.6)
```

As a rule of thumb: use `mapview` or `tmap` for exploration, `leaflet` for
polished outputs up to ~10,000 features, `leafgl` for up to ~500,000 features,
and `mapgl` for anything larger or requiring server-side tile rendering.

---

# Shiny integration

All of the packages covered here work inside Shiny applications. The most common
pattern with `leaflet` is:

- `leafletOutput()` in the UI to reserve space for the map.
- `renderLeaflet()` in the server to create the initial map.
- `leafletProxy()` to update the map reactively without redrawing it from scratch
  (e.g. to update a layer when a slider changes, without resetting the zoom level).

A minimal example:

```{r eval=FALSE}
library(shiny)

ui <- fluidPage(
  selectInput("year", "Year:", choices = 1985:2019, selected = 2019),
  leafletOutput("map", height = 500)
)

server <- function(input, output, session) {

  smb_all <- read_csv("ftp://ftp.hafro.is/pub/data/csv/is_smb.csv")

  # Render the base map once
  output$map <- renderLeaflet({
    leaflet() |>
      addProviderTiles(providers$Esri.OceanBasemap) |>
      setView(lng = -19, lat = 65, zoom = 6)
  })

  # Update only the data layer when the year changes
  observeEvent(input$year, {
    dat <- smb_all |>
      filter(year == input$year) |>
      st_as_sf(coords = c("lon1", "lat1"), crs = 4326)

    leafletProxy("map") |>
      clearGroup("stations") |>
      addCircleMarkers(data        = dat,
                       radius      = 4,
                       color       = "steelblue",
                       fillOpacity = 0.7,
                       weight      = 0,
                       group       = "stations",
                       label       = ~as.character(year))
  })
}

shinyApp(ui, server)
```

For `mapview` and `tmap`, equivalent Shiny output functions are
`mapviewOutput()` / `renderMapview()` and `tmapOutput()` / `renderTmap()`.

---

# Summary

| Package | Best for | Key strength |
|---|---|---|
| `mapview` | Quick exploration | Zero setup, sf/terra-aware |
| `tmap` | Report maps (static + interactive) | Grammar-of-graphics, mode switch |
| `leaflet` | Polished interactive outputs | Maximum control, wide ecosystem |
| `ggiraph` | ggplot-based interactivity | Same code → static or interactive |
| `leafgl` | Large datasets (~500k features) | WebGL rendering, leaflet API |
| `mapgl` | Very large / tile-based datasets | MapLibre GL, no API key needed |

## Further reading

- [Leaflet for R](https://rstudio.github.io/leaflet/)
- [tmap: Thematic Maps in R](https://r-tmap.github.io/tmap/)
- [mapview documentation](https://r-spatial.github.io/mapview/)
- [Geocomputation with R — Making maps with R](https://r.geocompx.org/adv-map)
- [ggiraph documentation](https://davidgohel.github.io/ggiraph/)
- [leafgl documentation](https://r-spatial.github.io/leafgl/)
- [mapgl documentation](https://walker-data.com/mapgl/)
