---
title: "Interpolation"
execute:
  enabled: false
---

**Needed libraries**:

```{r setup}
library(tidyverse)
library(sf)
library(terra)
library(tidyterra)
library(patchwork)
library(gstat)
library(automap)
library(mgcv)
```

# What is interpolation?

Interpolation is the process of predicting the value of some variable at unsampled
locations, based on observations at sampled ones. In spatial analysis, it is most
commonly used to convert a set of point observations into a continuous surface —
typically a raster.

Every interpolation method is a model, and every model makes assumptions. Choosing
the right method depends on the nature of your data and what you know (or can
reasonably assume) about the underlying spatial process. The main methods covered
here, roughly in order of complexity, are:

| Method | Assumptions | Best for |
|---|---|---|
| Voronoi / proximity polygons | Value constant within nearest-neighbour region | A simple, assumption-free baseline |
| Nearest neighbour (NN) | Value determined solely by the closest point(s) | Categorical data or a quick check |
| Inverse distance weighting (IDW) | Closer observations are more similar; no explicit model | Smooth continuous surfaces, fast |
| Ordinary kriging | Spatial autocorrelation follows a stationary variogram model | When spatial structure can be modelled; provides uncertainty estimates |
| GAMs with spatial smooths | Spatial trend is a smooth function of coordinates | When you also want to include non-spatial covariates |

::: {.callout-important}
## Project your data before interpolating
All distance-based interpolation methods require a projected coordinate system (metres, not degrees). Use `st_transform(x, crs)` to reproject before fitting any model. See the [CRS chapter](crs.qmd#which-projection-to-use) for guidance on choosing a projection appropriate to your study area.
:::

---

# Data preparation

We will interpolate cod biomass from the 2018 Icelandic spring bottom trawl survey
(SMB). The data are loaded once here and reused across all sections.

```{r}
biol <- read_csv("./data/is_smb_biological.csv") %>%
  filter(species == "cod")

cod <- read_csv("./data/is_smb_stations.csv") %>%
  filter(year == 2018) %>%
  left_join(biol, by = "id") %>%
  mutate(kg = replace_na(kg, 0)) %>%
  dplyr::select(id, lon1, lat1, kg, n, duration) %>%
  st_as_sf(coords = c("lon1", "lat1"), crs = 4326) %>%
  st_transform(3395)   # World Mercator — projected, units in metres

iceland <- read_sf("./data/iceland_coastline.gpkg") %>%
  st_transform(3395)
```

A quick look at the distribution of the response variable. Fisheries catch data
is typically zero-inflated and strongly right-skewed, which will matter when
choosing how to model it.

```{r}
p1 <- ggplot(cod) +
  geom_sf(aes(size = kg), alpha = 0.3, colour = "red",
          show.legend = "point") +
  geom_sf(data = iceland, fill = "grey80", colour = NA) +
  scale_size_area(max_size = 8) +
  labs(size = "Cod (kg)", title = "Station biomass") +
  theme_bw()

p2 <- ggplot(cod, aes(x = kg)) +
  geom_histogram(bins = 40, fill = "steelblue", colour = "white") +
  scale_x_continuous(labels = scales::comma) +
  labs(x = "Cod (kg)", y = "Count", title = "Distribution of kg") +
  theme_bw()

p1 + p2
```

## Building a prediction grid and mask

All raster-based interpolation methods need a target grid (where to predict) and
a mask (where predictions are meaningful). We build these once and reuse them.

The mask is defined as the convex hull of the sampling locations, buffered by
50 km, with land removed. This avoids predicting in areas far outside the survey
coverage or on land.

```{r}
# Convex hull of stations, buffered, land removed
pts_combined <- st_combine(cod)   # MULTIPOINT, needed for st_convex_hull

survey_mask <- st_convex_hull(pts_combined) %>%
  st_buffer(50000) %>%            # 50 km buffer
  st_difference(st_union(iceland)) %>%
  st_as_sf()

ggplot() +
  geom_sf(data = survey_mask, fill = "lightblue", alpha = 0.5) +
  geom_sf(data = iceland, fill = "grey80") +
  geom_sf(data = cod, colour = "red", size = 0.8) +
  labs(title = "Survey mask (blue) and stations (red)") +
  theme_bw()
```

Now we create a 2 km resolution target `SpatRaster` using `terra`, and rasterise
the mask onto it:

```{r}
bb <- st_bbox(survey_mask)

target <- terra::rast(
  xmin = bb["xmin"], xmax = bb["xmax"],
  ymin = bb["ymin"], ymax = bb["ymax"],
  resolution = 2000,
  crs = st_crs(cod)$wkt
)

mask_rst <- terra::rasterize(terra::vect(survey_mask), target, field = 1)
```

::: {.callout-note}
## Exercise

The 2 km resolution is a reasonable starting point for this survey, but it is
somewhat arbitrary. Repeat the data preparation with resolutions of 5 km and 10 km.
How does the choice of resolution affect the visual appearance and computation time
of the interpolations below?
:::

---

# Voronoi (proximity) polygons

The simplest spatial interpolation: each prediction location is assigned the value
of its nearest observation. The result is a set of Voronoi polygons (also called
Thiessen polygons), where each polygon contains all points closer to its seed
station than to any other.

Voronoi polygons make no assumptions about spatial autocorrelation and require no
model fitting. They are a good baseline to compare other methods against, but
produce discontinuous (blocky) surfaces.

```{r}
vor <- st_voronoi(pts_combined) %>%
  st_collection_extract("POLYGON") %>%
  st_intersection(survey_mask) %>%   # Clip to the survey area
  st_as_sf() %>%
  st_join(cod)                        # Attach the cod data via spatial join

ggplot() +
  geom_sf(data = vor, aes(fill = kg), colour = NA) +
  geom_sf(data = iceland, fill = "grey70", colour = NA) +
  geom_sf(data = cod, colour = "red", size = 0.5) +
  scale_fill_viridis_c(trans = "log1p", name = "Cod (kg)") +
  labs(title = "Voronoi interpolation") +
  theme_bw()
```

We can rasterise the Voronoi polygons onto the target grid using `terra::rasterize()`:

```{r}
vor_rst <- terra::rasterize(terra::vect(vor), target,
                            field = "kg", fun = "mean")
vor_rst <- terra::mask(vor_rst, mask_rst)

ggplot() +
  geom_spatraster(data = vor_rst) +
  geom_sf(data = iceland, fill = "grey70", colour = NA) +
  scale_fill_viridis_c(trans = "log1p", name = "Cod (kg)", na.value = NA) +
  labs(title = "Voronoi — rasterised to 2 km grid") +
  theme_bw()
```

---

# Inverse distance weighting (IDW)

In IDW, the predicted value at each location is a weighted average of all
observations. The weight assigned to each observation decays with distance,
controlled by the power parameter *p* (`idp` in `gstat`):

$$\hat{z}(s_0) = \frac{\sum_{i=1}^{n} w_i \, z(s_i)}{\sum_{i=1}^{n} w_i}, \quad w_i = \frac{1}{d(s_0, s_i)^p}$$

- A higher *p* gives more weight to nearby observations (more local, spikier surface).
- *p* = 0 gives equal weight to all observations everywhere (predicts the global mean).
- *p* = 2.5 is the conventional default.

`gstat::idw()` takes an *sf* object and a target `SpatRaster` directly:

```{r}
# Default p = 2 (idp = 2)
idw_p2 <- gstat::idw(kg ~ 1, locations = cod, newdata = target, idp = 2)
idw_p2 <- terra::mask(idw_p2[["var1.pred"]], mask_rst)

# Higher power — more local influence
idw_p5 <- gstat::idw(kg ~ 1, locations = cod, newdata = target, idp = 5)
idw_p5 <- terra::mask(idw_p5[["var1.pred"]], mask_rst)
```

```{r}
p_idw2 <- ggplot() +
  geom_spatraster(data = idw_p2) +
  geom_sf(data = iceland, fill = "grey70", colour = NA) +
  scale_fill_viridis_c(trans = "log1p", name = "Cod (kg)", na.value = NA) +
  labs(title = "IDW  (p = 2)") +
  theme_bw()

p_idw5 <- ggplot() +
  geom_spatraster(data = idw_p5) +
  geom_sf(data = iceland, fill = "grey70", colour = NA) +
  scale_fill_viridis_c(trans = "log1p", name = "Cod (kg)", na.value = NA) +
  labs(title = "IDW  (p = 5)") +
  theme_bw()

p_idw2 + p_idw5
```

The higher power (*p* = 5) produces a more pitted surface, with bull's-eye
patterns around individual stations. Notice how IDW never predicts values outside
the range of the observations, and always produces a minimum at sample locations
when surrounding values are higher — a known artefact.

IDW does not provide prediction uncertainty estimates.

::: {.callout-note}
## Exercise

The optimal *p* value can be selected by leave-one-out cross-validation (LOOCV):
for each station, remove it, predict its value from the remaining stations, and
compare the prediction to the observed value.

Use a `for` loop (or `purrr::map()`) over a grid of *p* values from 0.5 to 6 in
steps of 0.5. For each *p*, compute the root mean square error (RMSE) of the
LOOCV predictions. Which *p* minimises RMSE? Does it matter whether you compute
RMSE on raw or log-transformed values?

*Hint:* `gstat::idw()` with `nmax = nrow(cod) - 1` and `newdata = cod` gives
leave-one-out predictions in one call.
:::

---

# Ordinary kriging

Kriging is a family of geostatistical interpolation methods. Unlike IDW, kriging
derives the weights from a fitted model of the spatial autocorrelation structure
of the data — the *variogram*. This makes kriging a **Best Linear Unbiased
Predictor (BLUP)**: optimal in the sense of minimising prediction variance, given
the variogram model is correct. A key advantage over IDW is that kriging also
provides a **prediction variance** (uncertainty surface) alongside the predicted
values.

Ordinary kriging (OK) assumes:

1. The variable has a constant but unknown mean across the study area
   (*stationarity*).
2. The spatial correlation between two locations depends only on the distance
   (and possibly direction) between them, not on their absolute position
   (*second-order stationarity*).

For strongly skewed data like catch biomass, it is common to work in log space and
back-transform the predictions.

## Step 1 — Explore the data distribution

```{r}
ggplot(cod, aes(x = kg)) +
  geom_histogram(bins = 40, fill = "steelblue", colour = "white") +
  labs(x = "Cod biomass (kg)", y = "Count",
       title = "Raw distribution — strongly right-skewed") +
  theme_bw()

ggplot(cod, aes(x = log1p(kg))) +
  geom_histogram(bins = 40, fill = "steelblue", colour = "white") +
  labs(x = "log(1 + Cod biomass)", y = "Count",
       title = "Log(1+x) transformed — more symmetric") +
  theme_bw()
```

The log transformation makes the distribution more symmetric and closer to the
normality assumption that underlies kriging theory.

## Step 2 — Compute the experimental variogram

The variogram summarises how dissimilarity between pairs of observations increases
with the distance separating them. For a variable *Z* at locations *s*:

$$\hat{\gamma}(h) = \frac{1}{2|N(h)|} \sum_{(i,j) \in N(h)} [Z(s_i) - Z(s_j)]^2$$

where *N(h)* is the set of all pairs separated by lag distance *h*.

```{r}
# log1p transform before computing the variogram
vg_exp <- gstat::variogram(log1p(kg) ~ 1, data = cod)

plot(vg_exp,
     main = "Experimental variogram  —  log(1 + cod kg)")
```

Key features to look for:

- **Nugget**: the semivariance at distance ≈ 0. Represents measurement error and
  spatial variation at scales smaller than the sampling interval.
- **Sill**: the semivariance at which the variogram levels off. Roughly the
  variance of the data.
- **Range**: the distance at which the variogram reaches the sill. Beyond this
  distance, observations are no longer spatially autocorrelated.

## Step 3 — Fit a variogram model

A parametric model must be fitted to the experimental variogram before kriging.
`automap::autofitVariogram()` automates the model selection and fitting, trying
several common models and returning the best fit. This is a good starting point,
though manual fitting (as shown below) gives more control.

```{r}
# Automated fitting — useful as a starting point
vg_auto <- automap::autofitVariogram(log1p(kg) ~ 1, input_data = cod)
plot(vg_auto)
```

Manual fitting, using the experimental variogram as a guide:

```{r}
# Initial model: exponential with estimated nugget, sill and range
vg_model <- gstat::vgm(
  model  = "Exp",   # Exponential model
  psill  = 1.8,     # Partial sill (sill - nugget)
  range  = 200000,  # Range (metres)
  nugget = 0.3      # Nugget
)

# Fit to the experimental variogram using weighted least squares
vg_fit <- gstat::fit.variogram(vg_exp, model = vg_model)

plot(vg_exp, vg_fit,
     main = "Fitted variogram model (exponential)")
print(vg_fit)
```

Common variogram models available in `gstat`:

| Model | Shape | Use |
|---|---|---|
| `"Exp"` | Exponential | Asymptotically reaches sill; no true range |
| `"Sph"` | Spherical | Reaches sill at finite range; most common |
| `"Gau"` | Gaussian | Very smooth; often over-smooths |
| `"Mat"` | Matérn | Flexible; controlled by smoothness parameter *κ* |

::: {.callout-tip}
## Choosing a variogram model

There is no universally correct choice. The exponential and spherical models are
the most widely used. The key is that the fitted model should match the shape of
the experimental variogram well, particularly at short distances (the nugget and
the initial rise) where kriging weights are most sensitive.
:::

## Step 4 — Krige

`gstat::krige()` takes the *sf* data, the fitted variogram, and the target
`SpatRaster`. It returns a raster with two layers: `var1.pred` (predictions) and
`var1.var` (prediction variance).

```{r}
kriged <- gstat::krige(
  formula   = log1p(kg) ~ 1,
  locations = cod,
  newdata   = target,
  model     = vg_fit
)

# Extract and back-transform the prediction layer
pred_rst  <- terra::mask(expm1(kriged[["var1.pred"]]),  mask_rst)

# The variance is in log space — useful for uncertainty mapping
# but back-transformation is non-trivial; show it in log space
var_rst <- terra::mask(kriged[["var1.var"]], mask_rst)
names(pred_rst) <- "Predicted cod (kg)"
names(var_rst)  <- "Kriging variance (log scale)"
```

```{r}
p_krige <- ggplot() +
  geom_spatraster(data = pred_rst) +
  geom_sf(data = iceland, fill = "grey70", colour = NA) +
  geom_sf(data = cod, colour = "white", size = 0.4, alpha = 0.6) +
  scale_fill_viridis_c(trans  = "log1p",
                       name   = "Cod (kg)",
                       na.value = NA) +
  labs(title = "Ordinary kriging — predicted biomass") +
  theme_bw()

p_var <- ggplot() +
  geom_spatraster(data = var_rst) +
  geom_sf(data = iceland, fill = "grey70", colour = NA) +
  geom_sf(data = cod, colour = "white", size = 0.4, alpha = 0.6) +
  scale_fill_viridis_c(name = "Variance\n(log scale)",
                       na.value = NA,
                       option = "magma") +
  labs(title = "Kriging prediction variance",
       subtitle = "Higher = less certain") +
  theme_bw()

p_krige + p_var
```

Notice that prediction variance is lowest near sampling stations and increases
in areas with sparse coverage. This is a key advantage of kriging over IDW: it
makes its uncertainty explicit.

::: {.callout-note}
## Exercise

1. The variogram model above uses an exponential function. Refit the variogram
   using a spherical (`"Sph"`) and a Matérn (`"Mat"`) model. Compare the fitted
   curves visually and compare the resulting kriged surfaces. How sensitive are
   the predictions to the choice of variogram model?

2. Compute the 95% prediction interval at each cell. Under the assumption of
   log-normal errors, the interval on the original scale is approximately
   `[expm1(pred - 2*sqrt(var)), expm1(pred + 2*sqrt(var))]`. Map the width
   of this interval. Where is uncertainty greatest and why?
:::

## Universal kriging (kriging with external drift)

Ordinary kriging assumes a constant spatial mean. When you have a covariate that
explains part of the spatial trend — such as bottom depth explaining cod
distribution — **universal kriging** (also called kriging with external drift, KED)
incorporates it into the model. The formula changes from `log1p(kg) ~ 1` (constant
mean) to `log1p(kg) ~ depth` (mean varies as a linear function of depth):

```{r eval=FALSE}
# Example — requires a depth raster resampled to the same grid as `target`
depth_at_stations <- terra::extract(depth_rst, cod)
cod$depth <- depth_at_stations[, 2]

vg_uk <- gstat::variogram(log1p(kg) ~ depth, data = cod)
vg_uk_fit <- gstat::fit.variogram(vg_uk, gstat::vgm("Exp"))

kriged_uk <- gstat::krige(
  formula   = log1p(kg) ~ depth,
  locations = cod,
  newdata   = target,   # target must also contain a `depth` layer
  model     = vg_uk_fit
)
```

Universal kriging can substantially improve predictions when a good covariate is
available, but it requires the covariate to be known everywhere in the prediction
domain (i.e. on the target grid), not just at the sample locations.

---

# GAMs with spatial smooths

Generalised Additive Models (GAMs) offer a regression-based alternative to
kriging that fits naturally into the tidyverse and `mgcv` modelling workflow. A
2D spatial smooth `s(X, Y)` over projected coordinates acts as a nonparametric
spatial trend surface. Because GAMs are regression models, adding non-spatial
covariates (depth, temperature, vessel effects) is straightforward.

The Tweedie distribution (`family = tw()`) is a natural choice for catch biomass:
it handles the mixture of zeros and positive continuous values without requiring
a two-stage delta model.

## Fit the model

```{r}
# Extract projected coordinates as columns — mgcv needs them as plain variables
cod_xy <- bind_cols(cod, as_tibble(st_coordinates(cod)))

gam_mod <- mgcv::gam(
  kg ~ s(X, Y, k = 60),   # 2D spatial smooth; k controls maximum complexity
  data   = cod_xy,
  family = mgcv::tw(link = "log"),
  method = "REML"          # Restricted maximum likelihood for smooth selection
)

summary(gam_mod)
```

`k = 60` sets an upper bound on the degrees of freedom of the smooth. It should
be large enough not to be a binding constraint — check with `mgcv::k.check(gam_mod)`.

## Predict onto the grid

```{r}
# Get coordinates for all non-masked cells
cell_idx  <- which(!is.na(terra::values(mask_rst)))
pred_coords <- terra::xyFromCell(target, cell_idx) %>%
  as_tibble() %>%
  rename(X = x, Y = y)

# Predict with standard errors (for uncertainty)
pred_gam <- predict(gam_mod,
                    newdata = pred_coords,
                    type    = "response",
                    se.fit  = TRUE)

# Write predictions back into a raster
gam_pred_rst  <- mask_rst
gam_se_rst    <- mask_rst

terra::values(gam_pred_rst)[cell_idx] <- pred_gam$fit
terra::values(gam_se_rst)[cell_idx]   <- pred_gam$se.fit

names(gam_pred_rst) <- "GAM predicted (kg)"
names(gam_se_rst)   <- "GAM SE (response scale)"
```

```{r}
p_gam <- ggplot() +
  geom_spatraster(data = gam_pred_rst) +
  geom_sf(data = iceland, fill = "grey70", colour = NA) +
  geom_sf(data = cod, colour = "white", size = 0.4, alpha = 0.6) +
  scale_fill_viridis_c(trans = "log1p", name = "Cod (kg)",
                       na.value = NA) +
  labs(title = "GAM prediction  (Tweedie, s(X,Y))") +
  theme_bw()

p_gam_se <- ggplot() +
  geom_spatraster(data = gam_se_rst) +
  geom_sf(data = iceland, fill = "grey70", colour = NA) +
  scale_fill_viridis_c(name = "SE (kg)", na.value = NA, option = "magma") +
  labs(title = "GAM prediction standard error") +
  theme_bw()

p_gam + p_gam_se
```

::: {.callout-note}
## Exercise

1. Examine the model summary. What proportion of deviance is explained? Is the
   effective degrees of freedom of `s(X, Y)` close to the `k` limit? If so,
   increase `k` and refit.

2. Add bottom depth as a covariate: `kg ~ s(X, Y, k = 60) + s(depth, k = 10)`.
   You will need to extract depth values at each station from a depth raster
   (see the `rasters.qmd` chapter). Compare AIC between the two models. Does
   depth improve the fit?

3. Check residual spatial autocorrelation using a variogram of the GAM residuals.
   If autocorrelation remains, what does that suggest about the model?
:::

---

# Method comparison

A side-by-side comparison of all methods using the same colour scale helps
highlight where they agree and disagree. Large differences between methods in a
given area are a signal of high uncertainty, regardless of what any single method
reports.

```{r}
# Stack all prediction rasters with consistent names
names(vor_rst)      <- "Voronoi"
names(idw_p2)       <- "IDW (p=2)"
names(pred_rst)     <- "Kriging"
names(gam_pred_rst) <- "GAM"

all_preds <- c(vor_rst, idw_p2, pred_rst, gam_pred_rst)

ggplot() +
  geom_spatraster(data = all_preds) +
  geom_sf(data = iceland, fill = "grey70", colour = NA) +
  scale_fill_viridis_c(trans    = "log1p",
                       name     = "Cod (kg)",
                       na.value = NA) +
  facet_wrap(~lyr) +
  labs(title = "Interpolation method comparison  —  cod biomass 2018") +
  theme_bw() +
  theme(axis.text = element_blank(), axis.ticks = element_blank())
```

---

# Cross-validation

Visual comparison tells you how methods *differ*; cross-validation tells you how
they *perform*. Leave-one-out cross-validation (LOOCV) withholds each observation
in turn and predicts it from the rest. The prediction errors can then be summarised
as RMSE or mean absolute error (MAE).

The `gstat` package computes LOOCV predictions efficiently for IDW and kriging via
`gstat::krige.cv()`:

```{r}
# LOOCV for kriging
cv_krige <- gstat::krige.cv(
  formula   = log1p(kg) ~ 1,
  locations = cod,
  model     = vg_fit
)

# Back-transform and compute errors on the original scale
cv_krige <- cv_krige %>%
  mutate(
    pred_bt  = expm1(var1.pred),
    error    = pred_bt - kg,
    abs_err  = abs(error)
  )

rmse_krige <- sqrt(mean(cv_krige$error^2))
mae_krige  <- mean(cv_krige$abs_err)

cat("Kriging LOOCV  —  RMSE:", round(rmse_krige, 1),
    "  MAE:", round(mae_krige, 1), "\n")

# Map the residuals — spatial patterns indicate model misspecification
ggplot() +
  geom_sf(data = cv_krige,
          aes(colour = error, size = abs_err),
          alpha = 0.7) +
  geom_sf(data = iceland, fill = "grey80", colour = NA) +
  scale_colour_gradient2(low  = "blue", mid = "white", high = "red",
                         name = "Error (kg)") +
  scale_size_area(max_size = 8, name = "|Error| (kg)") +
  labs(title = "Kriging LOOCV residuals",
       subtitle = "Red = over-prediction, blue = under-prediction") +
  theme_bw()
```

::: {.callout-note}
## Exercise

Compute LOOCV RMSE for IDW with `p = 2` by running `gstat::krige.cv()` with a
`gstat` object that has no variogram model (i.e. pure IDW). Compare the RMSE
values for Voronoi, IDW (*p* = 2), kriging, and the GAM. Which method performs
best on this dataset? Does the "best" method also produce the most visually
plausible map?
:::

---

# Summary

- **Voronoi polygons** are the simplest possible interpolator and a useful
  baseline. Use them to sanity-check other results.
- **IDW** is fast and requires no model fitting. The power parameter *p* controls
  the degree of localisation. Use LOOCV to choose *p*.
- **Ordinary kriging** is the gold standard for continuous spatial prediction when
  the stationarity assumption is reasonable. It uniquely provides a spatial
  uncertainty estimate alongside the prediction.
- **Universal kriging** extends OK by incorporating external covariates (depth,
  temperature) into the spatial mean structure.
- **GAMs** with spatial smooths are a flexible regression-based alternative that
  integrate naturally with other predictors and are available in a familiar
  modelling framework (`mgcv`).
- Always **cross-validate** your interpolation. A method that looks smooth and
  plausible can still perform poorly out of sample.
- Working in **log space** (then back-transforming) is almost always preferable
  for skewed positive data such as fisheries catch biomass.

## Further reading

- Pebesma & Bivand (2023). *Spatial Data Science with R and the sf and stars packages*. <https://r-spatial.org/book/>
- Wood, S.N. (2017). *Generalized Additive Models: An Introduction with R* (2nd ed.). Chapman & Hall.
- Cressie, N. (1993). *Statistics for Spatial Data*. Wiley.
- `gstat` vignette: <https://cran.r-project.org/web/packages/gstat/vignettes/gstat.pdf>
- `automap` package: <https://cran.r-project.org/package=automap>
