---
title: "OSPAR Otter Trawl Fishing Intensity (2015)"
execute:
  enabled: true
---

# Background

The OSPAR Commission coordinates data collection on fishing intensity across
the Northeast Atlantic under the Joint Assessment and Monitoring Programme
(JAMP). The dataset used here — **OSPAR Otter Trawl Bottom Fishing Intensity
2015** — is a gridded product derived from Vessel Monitoring System (VMS) and
logbook data submitted through the ICES VMS Data Call. Several participants on
this course contribute to that data call directly.

Each grid cell follows the **C-square** coding system: a hierarchical spatial
reference that encodes location and resolution in a single string
(`"1400:390:130:1"`). At the resolution used here, cells are approximately
0.05° × 0.05°. The dataset covers the full OSPAR maritime area (Northeast
Atlantic, 36–82°N) with 91,190 cells.

### Key variables

| Variable | Description |
|---|---|
| `SurfSAR` | Surface Swept Area Ratio — total trawled area ÷ cell area |
| `SubsurfSAR` | Subsurface (penetrating) Swept Area Ratio |
| `FishingH` | Total fishing hours in the cell |
| `KWfishingH` | kW-fishing hours (effort weighted by vessel power) |
| `totweight` | Total catch weight (kg) |
| `totvalue` | Total catch value (€) — 41% missing |
| `mid_lon`, `mid_lat` | Cell centroid coordinates |

A value of **−9** is used throughout as a missing data sentinel and must be
recoded to `NA` before any analysis.

---

# Getting started

## Libraries

```{r setup}
library(tidyverse)
library(sf)
library(rnaturalearth)
library(rnaturalearthdata)
library(patchwork)
```

## Data

```{r data}
# OSPAR intensity grid — replace the -9 sentinel with NA
ospar <- read_sf("data/OSPAR_intensity_Otter_2015.gpkg") |>
  mutate(across(SubsurfSAR:FishingH, \(x) if_else(x == -9, NA_real_, x)))

# ICES ecoregions — st_make_valid() needed for a geometry issue (see §4)
ices_er <- read_sf("data/ices_ecoregions.gpkg") |>
  select(ecoregion) |>
  st_make_valid()

# Land background
land <- ne_countries(scale = "medium", returnclass = "sf")
```

---

# Data quality

The −9 sentinel is not applied uniformly across variables:

```{r missing}
ospar |>
  st_drop_geometry() |>
  summarise(across(SubsurfSAR:FishingH,
                   \(x) scales::percent(mean(is.na(x)), accuracy = 0.1))) |>
  pivot_longer(everything(), names_to = "variable", values_to = "pct_missing")
```

`totvalue` has ~41% missing values and should be used cautiously. All other
variables are ~2.7% missing — consistent with cells where the gear type or
data quality flag prevented reporting.

---

# Mapping fishing intensity

## Performance note

The OSPAR dataset is provided as 91,190 polygons. Plotting them with `geom_sf()`
renders every polygon outline and is slow (10–30 seconds on most machines). A
much faster approach is to use the pre-computed centroid columns `mid_lon` and
`mid_lat` with `geom_tile()`, treating the dataset as a raster grid. We keep
the polygon geometry only for spatial operations (§4).

```{r map-setup}
# Data frame version — no geometry, fast to plot
ospar_df <- ospar |> st_drop_geometry()
```

## Surface swept area ratio (SurfSAR)

SAR is the standard OSPAR metric for trawling disturbance: values > 1 mean the
seabed was trawled more than once during the year. A log(1 + x) scale is
appropriate given the strongly right-skewed distribution.

```{r map-surfsar}
ospar_df |>
  filter(!is.na(SurfSAR)) |>
  ggplot() +
  geom_tile(aes(x = mid_lon, y = mid_lat, fill = SurfSAR)) +
  geom_sf(data = land, fill = "grey75", colour = "white", linewidth = 0.2,
          inherit.aes = FALSE) +
  scale_fill_viridis_c(option = "inferno", direction = -1,
                       transform = "log1p",
                       name = "Surface SAR\n(log scale)") +
  coord_sf(xlim = c(-42, 51), ylim = c(36, 82)) +
  labs(x = NULL, y = NULL,
       title = "OSPAR otter trawl surface swept area ratio, 2015") +
  theme_bw()
```

## Surface vs subsurface SAR

Subsurface SAR measures gear penetration into the sediment — ecologically
important for benthic habitat disturbance assessment. Comparing the two
reveals where demersal trawling causes deeper seabed impact.

```{r map-sar-comparison}
make_sar_map <- function(data, var, title) {
  data |>
    filter(!is.na(.data[[var]])) |>
    ggplot() +
    geom_tile(aes(x = mid_lon, y = mid_lat, fill = .data[[var]])) +
    geom_sf(data = land, fill = "grey75", colour = "white", linewidth = 0.2,
            inherit.aes = FALSE) +
    scale_fill_viridis_c(option = "inferno", direction = -1,
                         transform = "log1p", name = "SAR\n(log scale)") +
    coord_sf(xlim = c(-15, 15), ylim = c(47, 62)) +  # North Sea / Celtic Seas
    labs(x = NULL, y = NULL, title = title) +
    theme_bw() +
    theme(legend.position = "bottom")
}

make_sar_map(ospar_df, "SurfSAR",    "Surface SAR") +
make_sar_map(ospar_df, "SubsurfSAR", "Subsurface SAR")
```

## Fishing hours and effort

`FishingH` and `KWfishingH` are effort metrics. The kW-weighted version
up-weights larger vessels and is a better proxy for fishing capacity. Plotting
both side-by-side highlights where small versus large vessels dominate:

```{r map-effort}
p_hours <- ospar_df |>
  filter(!is.na(FishingH)) |>
  ggplot() +
  geom_tile(aes(x = mid_lon, y = mid_lat, fill = FishingH)) +
  geom_sf(data = land, fill = "grey75", colour = "white", linewidth = 0.2,
          inherit.aes = FALSE) +
  scale_fill_viridis_c(option = "mako", direction = -1,
                       transform = "log1p", name = "Hours\n(log)") +
  coord_sf(xlim = c(-15, 15), ylim = c(47, 62)) +
  labs(x = NULL, y = NULL, title = "Fishing hours") +
  theme_bw() + theme(legend.position = "bottom")

p_kw <- ospar_df |>
  filter(!is.na(KWfishingH)) |>
  ggplot() +
  geom_tile(aes(x = mid_lon, y = mid_lat, fill = KWfishingH)) +
  geom_sf(data = land, fill = "grey75", colour = "white", linewidth = 0.2,
          inherit.aes = FALSE) +
  scale_fill_viridis_c(option = "mako", direction = -1,
                       transform = "log1p", name = "kW·h\n(log)") +
  coord_sf(xlim = c(-15, 15), ylim = c(47, 62)) +
  labs(x = NULL, y = NULL, title = "kW-fishing hours") +
  theme_bw() + theme(legend.position = "bottom")

p_hours + p_kw
```

---

# Aggregation by ICES ecoregion

Spatial joining 91,190 grid cells against polygon boundaries is the key
operation here. The ICES ecoregion polygons contain a minor geometry validity
issue that triggers an S2 error. `st_make_valid()` resolves it cleanly and
was already applied when loading the data above.

```{r ecoregion-join}
# The join takes ~2–3 seconds on a typical laptop.
# sf_use_s2(FALSE) is required: the ICES ecoregion polygons have a topology
# that S2 spherical geometry does not accept (even after st_make_valid()).
sf_use_s2(FALSE)
by_region <- ospar |>
  st_join(ices_er) |>
  st_drop_geometry() |>
  filter(!is.na(ecoregion)) |>
  group_by(ecoregion) |>
  summarise(
    total_FishingH   = sum(FishingH,    na.rm = TRUE),
    mean_SurfSAR     = mean(SurfSAR,    na.rm = TRUE),
    mean_SubsurfSAR  = mean(SubsurfSAR, na.rm = TRUE),
    total_SurfSA_km2 = sum(SurfSA,     na.rm = TRUE),
    total_weight_kt  = sum(totweight,   na.rm = TRUE) / 1e6,
    .groups = "drop"
  ) |>
  arrange(desc(total_FishingH))
sf_use_s2(TRUE)

by_region
```

## Bar chart — total fishing hours by ecoregion

```{r ecoregion-bar}
by_region |>
  mutate(ecoregion = fct_reorder(ecoregion, total_FishingH)) |>
  ggplot(aes(x = total_FishingH / 1000, y = ecoregion)) +
  geom_col(fill = "steelblue") +
  labs(x = "Total fishing hours (thousands)", y = NULL,
       title = "OSPAR otter trawl effort by ICES ecoregion, 2015") +
  theme_bw()
```

## Choropleth — mean SurfSAR per ecoregion

Joining the aggregated table back to the ecoregion polygons gives a choropleth.
`mean_SurfSAR` is the average SAR over cells that were actually trawled, so it
measures **intensity where fishing occurred** — not what fraction of the
ecoregion was affected.

```{r ecoregion-choropleth}
ices_er_summary <- ices_er |>
  left_join(by_region, by = "ecoregion")

ggplot() +
  geom_sf(data = ices_er_summary, aes(fill = mean_SurfSAR), colour = "white") +
  scale_fill_viridis_c(option = "inferno", direction = -1,
                       name = "Mean\nSurface SAR",
                       na.value = "grey90") +
  coord_sf(xlim = c(-42, 51), ylim = c(36, 82)) +
  labs(x = NULL, y = NULL,
       title = "Mean otter trawl surface SAR by ICES ecoregion, 2015",
       subtitle = "Average over trawled cells only") +
  theme_bw()
```

## Area-normalised SAR per ecoregion

`mean_SurfSAR` can be misleading: a region where fishing is extremely intense
in a small corner will look similar to one where the same intensity is spread
across the whole area. A more complete picture requires normalising by the total
ecoregion area rather than the number of trawled cells.

The **region-wide SAR** is:

$$\text{region SAR} = \frac{\sum \text{SurfSA (km}^2\text{)}}{\text{ecoregion area (km}^2\text{)}}$$

where `SurfSA` is the actual swept area (km²) per cell. Ecoregion areas are
computed from the polygon geometries using `st_area()` — the `area_km2` column
stored in the file is inconsistent with the geometries and should not be used.

```{r region-sar}
# Ecoregion areas from geometry (authoritative).
# The area_km2 column stored in the file is inconsistent with the polygon
# geometries and should not be used.
ecoregion_areas <- ices_er |>
  mutate(area_km2 = as.numeric(st_area(geom)) / 1e6) |>
  st_drop_geometry() |>
  select(ecoregion, area_km2)

# by_region already contains total_SurfSA_km2 from the join above
region_sar <- by_region |>
  left_join(ecoregion_areas, by = "ecoregion") |>
  mutate(region_SAR = total_SurfSA_km2 / area_km2) |>
  arrange(desc(region_SAR))

region_sar |> select(ecoregion, mean_SurfSAR, total_SurfSA_km2, area_km2, region_SAR)
```

The contrast between the two metrics is stark:

```{r region-sar-comparison}
region_sar |>
  select(ecoregion, mean_SurfSAR, region_SAR) |>
  pivot_longer(cols = c(mean_SurfSAR, region_SAR),
               names_to = "metric", values_to = "value") |>
  mutate(
    ecoregion = fct_reorder(ecoregion, value, .fun = max),
    metric = recode(metric,
                    mean_SurfSAR = "Mean SAR (trawled cells)",
                    region_SAR   = "Region SAR (whole ecoregion)")
  ) |>
  ggplot(aes(x = value, y = ecoregion, colour = metric, shape = metric)) +
  geom_point(size = 3) +
  scale_colour_manual(values = c("Mean SAR (trawled cells)"  = "coral",
                                 "Region SAR (whole ecoregion)" = "steelblue"),
                      name = NULL) +
  scale_shape_manual(values = c("Mean SAR (trawled cells)"  = 16,
                                "Region SAR (whole ecoregion)" = 17),
                     name = NULL) +
  labs(x = "SAR", y = NULL,
       title = "Trawl intensity vs areal coverage by ecoregion",
       subtitle = "Gap between metrics indicates spatially concentrated fishing") +
  theme_bw() +
  theme(legend.position = "bottom")
```

The gap between the two metrics is informative. The **Bay of Biscay** has the
highest mean SAR per trawled cell (2.8 — each cell is swept nearly three times
on average) yet covers only 54% of the ecoregion area, indicating highly
concentrated fishing. The **Celtic Seas** and **Greater North Sea** have a
smaller gap: nearly the entire area (99% and 95% respectively) was swept at
least once in 2015, making them the most extensively impacted ecoregions
regardless of metric.

```{r region-sar-choropleth}
ices_er_sar <- ices_er |>
  left_join(region_sar, by = "ecoregion")

ggplot() +
  geom_sf(data = ices_er_sar, aes(fill = region_SAR), colour = "white") +
  scale_fill_viridis_c(option = "inferno", direction = -1,
                       name = "Region SAR\n(area-normalised)",
                       na.value = "grey90",
                       labels = scales::percent_format(accuracy = 1)) +
  coord_sf(xlim = c(-42, 51), ylim = c(36, 82)) +
  labs(x = NULL, y = NULL,
       title = "Otter trawl surface SAR normalised by ecoregion area, 2015",
       subtitle = "Proportion of each ecoregion's seabed swept in 2015") +
  theme_bw()
```

## Surface vs subsurface disturbance by ecoregion

```{r ecoregion-sar-comparison}
by_region |>
  select(ecoregion, mean_SurfSAR, mean_SubsurfSAR) |>
  pivot_longer(cols = c(mean_SurfSAR, mean_SubsurfSAR),
               names_to = "type", values_to = "mean_SAR") |>
  mutate(
    ecoregion = fct_reorder(ecoregion, mean_SAR, .fun = max),
    type = recode(type,
                  mean_SurfSAR    = "Surface SAR",
                  mean_SubsurfSAR = "Subsurface SAR")
  ) |>
  ggplot(aes(x = mean_SAR, y = ecoregion, fill = type)) +
  geom_col(position = "dodge") +
  scale_fill_manual(values = c("Surface SAR"    = "steelblue",
                               "Subsurface SAR" = "coral"),
                    name = NULL) +
  labs(x = "Mean SAR", y = NULL,
       title = "Surface vs subsurface trawl disturbance by ecoregion, 2015") +
  theme_bw() +
  theme(legend.position = "bottom")
```

---

# Exercises

::: {.callout-note}
## Exercise 1 — Celtic Seas zoom

The Celtic Seas has the highest total fishing hours in the dataset. Filter the
OSPAR data to `mid_lon` between −15 and 0 and `mid_lat` between 47 and 57, and
map `SurfSAR` at this finer scale. Add `iceland_contours.gpkg` depth contours
as a reference layer. What is the relationship between depth and fishing
intensity?
:::

::: {.callout-note}
## Exercise 2 — Catch value vs effort

`totvalue` is missing for ~41% of cells. Investigate whether the missing values
are spatially structured (i.e., concentrated in specific regions) by mapping
`is.na(totvalue)`. Then, for cells with non-missing values, plot `totvalue`
against `KWfishingH` — do high-effort cells always generate high-value catches?
:::

::: {.callout-note}
## Exercise 3 — ORE/MSP overlap (advanced)

`ices_ecoregions.gpkg` and `nephrops_fu.gpkg` are both polygon layers that
define management boundaries. Use `st_intersection()` to clip the OSPAR data
to a specific Nephrops functional unit of your choice and compute the total
fishing effort (`FishingH`) and mean SAR within it. How might this information
be used in an environmental impact assessment for an offshore wind development?
:::
