---
title: "Species distribution model"
output:
  html_document:
    fig_height: 4
    fig_width: 8
    highlight: haddock
    theme: cerulean
    toc: yes
    toc_float: yes
---




## The data

For this exercise we will use as predictors the following datasets:

- Bottom depth
- Bottom slope
- Temperature


#### Silica
We will obtain data on near-bottom concentration of Silicate using the package **sdmpredictors**, which provides access to the Bio-ORACLE

```{r }

library(sdmpredictors)
options(sdmpredictors_datadir="./data_raw/")

xlim <- c(-28, -10)
ylim <- c(62.5, 67.5)

silica <- load_layers("BO2_silicateltmin_bdmean") %>%
  crop(extent(c(xlim, ylim)))

plot(silica, axes = TRUE, main = "Silica")


```
#### Depth
We will download bathymetry data from NOAA's ETOPO1 database using the *getNOAA.bathy()* function in the **marmap** package.  This package uses its own class but it is easy to convert it to a raster.

```{r }
xlim <- c(-28, -10)
ylim <- c(62.5, 67.5)

library(marmap)

setwd("./data_raw") # Because the data gets downloaded to your current directory

depth <-
  getNOAA.bathy(lon1 = xlim[1], lon2 = xlim[2],
                lat1 = ylim[1], lat2 = ylim[2],
                resolution = 1, keep=TRUE) %>% # Write the data into your hard drive
  as.raster()  # turn the object into a raster

setwd("..")

plot(depth, axes = TRUE, main = "Depth")


```
