---
title: "R and QGIS"
---
  
R and QGIS are complementary tools, and many practitioners use both in the same 
workflow. QGIS excels at interactive, exploratory work: visually inspecting data, 
digitising features, and quick one-off queries. R excels at reproducible, scripted 
analysis that can be run end-to-end by anyone with the same data. There is no 
need to choose between them.

This chapter covers three ways to connect the two: simple file exchange, calling 
QGIS processing algorithms directly from R, and accessing GRASS GIS through R.

---
  
# The shared foundation: GDAL
  
R and QGIS are both built on [GDAL](https://gdal.org/), the Geospatial Data 
Abstraction Library. This means they can read and write the same file formats, 
and a file written by one can always be opened by the other. You do not need any 
special integration layer to move data between them — a GeoPackage is a 
GeoPackage whether you created it in R or in QGIS.

```{r setup}
library(tidyverse)
library(sf)
library(terra)
```

---
  
# File exchange
  
The simplest and most reliable way to combine R and QGIS in a workflow is to 
write files from R that QGIS can open, and to read files prepared in QGIS back 
into R.

**GeoPackage** (`.gpkg`) is the recommended format for this. It is an open 
standard supported natively by both tools, stores multiple layers in a single 
file, and avoids the limitations of shapefiles. **GeoTIFF** (`.tif`) is the 
equivalent for rasters.


::: {.callout-tip}
## Organise your project folders clearly
A simple convention that works well: keep a `qgis/` folder for layers you create 
or edit interactively in QGIS, and an `output/` folder for layers produced by 
your R scripts. Reading from `qgis/` and writing to `output/` keeps the 
direction of data flow obvious.
:::
  
  ---
  
  # Calling QGIS algorithms from R: `qgisprocess`

The **qgisprocess** package (Dunnington et al.) lets you run QGIS processing 
algorithms directly from an R script, passing `sf` and `terra` objects as inputs 
and getting results back as R objects. This is the current recommended approach 
for tight R--QGIS integration.

It requires a local QGIS installation (version 3.14 or later), and it wraps the 
`qgis_process` command-line tool that QGIS provides. All providers available in 
your QGIS installation are accessible: native QGIS algorithms, GDAL, GRASS, and 
SAGA.



## Configuring qgisprocess

```{r}
install.packages("qgisprocess")

library(qgisprocess)

# Point to qgis_process in a Mac
options(qgisprocess.path = "/Applications/QGIS-4.0.1.app/Contents/MacOS/qgis_process")

# In Windows is a bat file commonly found at a location like this:
# C:\Program Files\QGIS 3.x\bin\qgis_process-qgis.bat

qgis_configure()

```

## Checking your setup

```{r}
# Verify that qgisprocess can find your QGIS installation
qgis_path()
qgis_version()
```

## Discovering available algorithms

```{r}
# List all available algorithms
qgis_algorithms()

# Search by keyword
qgis_algorithms() |>
  filter(grepl("buffer", algorithm, ignore.case = TRUE)) |>
  select(provider, algorithm, algorithm_title)

# Get the full documentation for one algorithm
qgis_show_help("native:buffer")
```

## A worked example: buffering survey stations

```{r}
# Load survey stations
stations <- read_sf("./data/is_smb_stations.csv") |>
  st_as_sf(coords=c("lon1","lat1"),crs=4326) |> 
  st_transform(3057)   # reproject for buffering

# Run the QGIS buffer algorithm
result <- qgis_run_algorithm(
  algorithm = "native:buffer",
  INPUT     = stations,
  DISTANCE  = 5000,    # 5 km buffer, in the CRS units (metres)
  SEGMENTS  = 16,
  DISSOLVE  = FALSE
)

# Convert the output back to an sf object
station_buffers <- sf::read_sf(qgis_extract_output(result, "OUTPUT"))

# Inspect
glimpse(station_buffers)

# Quick plot
ggplot() + 
  geom_sf(data=station_buffers |> st_geometry())
```

::: {.callout-note}
The `DISTANCE` parameter is always in the units of the input layer's CRS. 
Reproject to a projected CRS (metres) before buffering if your data are in 
degrees.
:::

## Another example: dissolving polygons by attribute

```{r}

qgis_show_help("native:dissolve")

ospar.sr <- read_sf("./data/ospar.gpkg") 

sf_use_s2(FALSE)
sq <- read_sf("./data/OSPAR_intensity_Otter_2015.gpkg") |> 
  st_join(ospar.sr) |> 
  select(subregion) |> 
  st_transform(3035)
sf_use_s2(TRUE)


result <- qgis_run_algorithm(
  algorithm = "native:dissolve",
  INPUT     = sq,
  FIELD     = "subregion"
)

sq_dissolved <- read_sf(qgis_extract_output(result, "OUTPUT"))

ggplot()+
  geom_sf(data=sq_dissolved, aes(fill=subregion))

```

## Chaining algorithms

`qgisprocess` results can be passed directly as inputs to the next algorithm 
without converting back to `sf` at each step:

```{r}
# Buffer, then dissolve, in one chain
buffered <- qgis_run_algorithm("native:buffer",
                                INPUT    = stations,
                                DISTANCE = 5000)

dissolved <- qgis_run_algorithm("native:dissolve",
                                 INPUT = qgis_extract_output(buffered, "OUTPUT"))

final <- read_sf(qgis_extract_output(dissolved, "OUTPUT"))
```

::: {.callout-warning}
## Package versions
**qgisprocess** stabilised on CRAN in 2023 and is the successor to the older 
`RQGIS` and `RQGIS3` packages, which are no longer maintained. If you encounter 
references to `RQGIS` in older tutorials or Stack Overflow answers, use 
`qgisprocess` instead.

Check the current version and documentation at: 
<https://r-spatial.github.io/qgisprocess/>
:::

---

# GRASS GIS from R: `rgrass`

[GRASS GIS](https://grass.osgeo.org/) is a powerful open-source GIS that 
includes algorithms not available in QGIS or R, particularly for terrain 
analysis (watershed delineation, viewshed, least-cost paths) and advanced raster 
processing. The **rgrass** package provides a direct interface between R and 
GRASS.

`qgisprocess` can also access GRASS algorithms if GRASS is installed alongside 
QGIS, so for many purposes you do not need `rgrass` directly. Use `rgrass` when 
you need fine-grained control over the GRASS environment, or when you are working 
with workflows built around GRASS natively.

```{r}
#| eval: false

# install.packages("rgrass")
library(rgrass)

# Initialise a GRASS session (requires GRASS to be installed)
initGRASS(gisBase  = "/usr/lib/grass82",   # path to your GRASS installation
          home     = tempdir(),
          override = TRUE)

# Write an sf object into the GRASS environment
write_VECT(my_sf, vname = "my_layer", v.in.ogr_flags = "overwrite")

# Run a GRASS command
execGRASS("v.buffer",
          input    = "my_layer",
          output   = "my_layer_buffered",
          distance = 5000)

# Read the result back into R as an sf object
result <- read_VECT("my_layer_buffered")
```

For most participants on this course, `qgisprocess` will be sufficient. The 
`rgrass` documentation is at <https://rsbivand.github.io/rgrass/>.

---

# Choosing the right approach

| Situation | Recommended approach |
|---|---|
| Move a processed layer to QGIS for visual checking | Write GeoPackage with `write_sf()` |
| Use a QGIS-prepared layer in R | Read GeoPackage with `read_sf()` |
| Run a specific QGIS algorithm inside a script | `qgisprocess` |
| Need GRASS algorithms (watersheds, terrain) | `qgisprocess` with GRASS provider, or `rgrass` |
| Building a fully reproducible pipeline | R throughout; QGIS for visual QC only |

::: {.callout-tip}
## A practical rule of thumb
Use QGIS for interactive exploration and visual checks. Use R for anything you 
need to run more than once, share with collaborators, or include in a report. 
File exchange via GeoPackage always works. `qgisprocess` is the best option when 
you specifically need a QGIS algorithm inside a scripted workflow.
:::

---

# Further reading

- [qgisprocess package documentation](https://r-spatial.github.io/qgisprocess/) — 
  full reference with vignettes and algorithm examples.
- [QGIS documentation: Processing framework](https://docs.qgis.org/latest/en/docs/user_manual/processing/index.html) — 
  describes all native algorithms accessible via `qgisprocess`.
- [rgrass documentation](https://rsbivand.github.io/rgrass/) — interface to GRASS GIS from R.
- [Geocomputation with R, Chapter 9: Bridges to GIS software](https://r.geocompx.org/gis) — 
  a detailed treatment of R--QGIS--GRASS--SAGA integration.
